Smed-pou4-2 regulates mechanosensory neuron regeneration and function in planarians

Version 1

Abstract

Abstract

POU4 homologs are involved in the development of sensory cell types across diverse species, including cnidarians, ascidians, and mammals. Whether these developmental regulators are reused during adult tissue maintenance and regeneration remains a fundamental question in regenerative biology. Here, we investigate the role of the Schmidtea mediterranea BRN3/POU4 homolog, Smed-pou4-2 (pou4-2), in the regeneration of mechanosensory neurons. We find that pou4-2 is regulated by the SoxB1 homolog, soxB1-2, and is expressed in a distinct population of ciliated sensory cells that detect water flow. Transcriptomic analysis of pou4-2-deficient planarians reveals enrichment for conserved genes associated with human auditory and vestibular function, suggesting that planarian rheosensory neurons share molecular features with mammalian inner ear hair cells. Expression of these conserved genes is abrogated by RNAi-mediated knockdown of pou4-2. To determine whether these transcriptional changes had functional consequences for mechanosensory neuron identity or behavior, we next assessed the impact of pou4-2 knockdown on sensory function. pou4-2 RNAi results in impaired mechanosensation in both uninjured and regenerating planarians. Together with the loss of terminal differentiation markers in mechanosensory neurons, these findings identify Smed-pou4-2 as a key regulator of mechanosensory neuron identity in planarians and support the idea that conserved sensory specification programs are redeployed during adult tissue regeneration.

Introduction

Most animals, including mammals, have a limited capacity for neuronal regeneration. In contrast, organisms like fish and salamanders can effectively regenerate neurons, and some invertebrates are capable of dramatic whole-body regeneration. The freshwater planarian Schmidtea mediterranea is among a handful of research organisms capable of restoring virtually any lost or damaged tissue and can regenerate entire animals from small body fragments (Goldstein and Srivastava, 2022; Ivankovic et al., 2019). S. mediterranea possesses a population of adult pluripotent stem cells called neoblasts, which proliferate and differentiate to replace all missing tissues (Baguñà, 2012; Newmark and Sánchez Alvarado, 2002; Reddien, 2018). This stem cell population is postulated to include a heterogeneous pluripotent pool poised to acquire lineage-specific cell fates as needed (Raz et al., 2021). One of the extraordinary properties of planarians is the capacity for constant neuronal turnover and regeneration of neuronal cell types, many of which are conserved with vertebrates (Brown and Pearson, 2017; Ross et al., 2017). However, we know very little about the molecular basis of neurogenesis or the signals that regulate neuronal turnover (Lee, 2023). Previous studies found that soxB1-2, a mammalian Sox1/2/3 homolog, regulates the regeneration of ectodermal cell type subsets in planarians, including many uncharacterized sensory neurons (Ross et al., 2018). One prominent population of soxB1-2-regulated sensory cells, organized in a striking dorsal stripe pattern, functions in mechanosensation and is marked by polycystic kidney disease-like homologs (Ross et al., 2024; Ross et al., 2018) (Fig. 1A). Here, we sought to investigate mechanisms downstream of soxB1-2 that are required to specify mechanosensory cells in the dorsal ciliated stripe.

Smed-pou4-2 is expressed in the ciliated stripes.

(A) Planarian cartoon illustrates the dorsal and peripheral cell patterns implicated in detecting mechanosensory stimulation in Schmidtea mediterranea. (B) Whole-mount in situ hybridization to pou4-2 revealed a stereotyped pattern depicted in (A), in the dorsal head tip, body periphery, dorsal ciliated stripe (dcs), and additional expression in the ventral nerve cords (vnc). The dashed line represents the cross-section plane shown in the bottom panel. Anterior is to the top. Scale bar = 200 μm. The cross-section image shows pou4-2 expression in the dorsal ciliated stripe (dcs), dorsal and ventral peripheral stripes (pcs), and ventral nerve cords (vnc). Scale bar = 200 µm. (C) Expression analysis of pou4-2. The worms were amputated pre-pharyngeally, and regenerates were fixed at the designated time points. The asterisk highlights the first reappearance of expression within the blastema. The blue arrows at 4 dpa denote the appearance of the anterior dorsal ciliated stripe pattern. Anterior is to the top; dpa, days post-amputation. Scale bar = 200 μm; n ≥ 3 worms tested, with all samples displaying similar expression patterns.

POU transcription factor family genes play key roles in the development and function of many neuronal subtypes. To date, dozens of POU genes have been identified in vertebrates and invertebrates, and their roles in the differentiation and survival of diverse neuronal subtypes have been characterized (Leyva-Diaz et al., 2020). In many species, Brn3/POU4 transcription factors play important roles in specifying and maintaining the identities of various cell populations in the developing peripheral sensory nervous system. A notable example is Nematostella vectensis NvPOU4, which is required to maintain and differentiate cnidocytes, a population of mechanosensing cells exclusive to the phylum Cnidaria (Tourniere et al., 2020). The homologous role of NvPOU4 in cnidarians suggests that the functional role of pou4 is ancient and conserved across distantly related phyla. Additionally, pou4 is part of a proneural regulatory cascade that produces epidermal sensory neurons in Ciona intestinalis; induction of ectopic pou4 expression in the developing epidermis of Ciona larvae converts epidermal cells to sensory neurons, resulting in a striking hyper-ciliated phenotype (Chen et al., 2011).

In mice, Pou4f3 is expressed exclusively in the inner ear sensory epithelia during embryonic development and is required for the survival of vestibular hair cells of the auditory system (Erkman et al., 1996; Xiang et al., 1997) - its targeted deletion results in impaired hearing and balance. Hair cells of the inner ear are crosslinked by stereocilia on their apical ends that function as mechanosensors, converting vibration-induced mechanical force into signals carried by auditory nerve fibers to the central nervous system (Goutman et al., 2015). Although a small number of hair cells differentiate in Pou4f3-/- mice, their failure to form stereociliary bundles leads to apoptosis (Xiang et al., 1998). Thus, Pou4 has conserved roles in the differentiation, maintenance, and survival of ciliated mechanosensory neurons. Unlike birds and fish, mammals lack the ability to regenerate hair cells after they are lost, resulting in permanent deafness (Edge and Chen, 2008). Recent studies show that POU4 can be used as a reprogramming co-factor to restore hair cells in mammals (Chen et al., 2021; Iyer et al., 2022). However, whether developmental regulators, like POU4, are reused during adult tissue maintenance and regeneration remains a fundamental question in regenerative biology (Seifert et al., 2023).

In S. mediterranea, a search for candidate planarian OCT4 homologs, a gatekeeper of pluripotency also known as POU5F1 in humans (Zeineddine et al., 2014), revealed six genes containing a POU-specific domain and a POU-homeodomain, and the two genes most similar to hPOU4F3 were named Smed-pou4-1 and Smed-pou4-2 (Önal et al., 2012b) (referred to as pou4-1 and pou4-2 hereon). pou4-1 (also referred to as pou4-like and pou4-like-1) was identified as downstream of COE (Cowles et al., 2014), a transcription factor required for neurogenesis widely conserved across metazoans (Demilly et al., 2011), and is responsible for maintaining proper neuronal architecture in the cephalic ganglia as well as photoreceptor pigmentation (Cowles et al., 2014). More recently, our lab and others observed robust sensory defects in pou4-2(RNAi) planarians (Elliott, 2016; McCubbin, 2022; Wang, 2019). Under normal conditions, S. mediterranea worms display a stereotyped behavior by shortening their bodies in response to vibrations and water currents (rheosensation) across their dorsal side. We found that pou4-2(RNAi) planarians failed to react to this sensory input, indicating a significant role for pou4-2 in mechanosensory neuron function (Elliott, 2016; Wang, 2019); however, its role in regenerative neurogenesis is not well understood. Here, we examined the function of pou4-2 in mechanosensory neuron regeneration.

In this study, we mapped the expression of pou4-2 and assessed its function through RNAi and RNA-seq. Loss of pou4-2 expression coincides with loss of mechanosensation, which is not restored in regenerated pou4-2(RNAi) planarians. Analysis of the pou4-2+ cell gene expression profile uncovered that many genes regulated by Pou4-2 activity are necessary for proper mechanosensory neuron function and are homologs of human genes involved in hair cell function and auditory perception. In many organisms, the proneural atonal genes function in the same gene regulatory network as pou4 (Leyva-Diaz et al., 2020). However, this relationship does not appear to be conserved in planarians. Our findings implicate pou4-2 in a regulatory cascade that specifies distinct sensory neuron populations. This study demonstrates that pou4-2 plays a key regulatory role in the differentiation, maintenance, and regeneration of ciliated mechanosensory neurons in planarians.

Results

Smed-pou4-2 is expressed in planarian mechanosensory neurons

Our previous work demonstrated the key role of soxB1-2 in sensory neuron differentiation and function of sensory neuron subclasses in the planarian Schmidtea mediterranea (Ross et al., 2018). A subset of soxB1-2-regulated genes are abundantly expressed in a discrete pattern called the dorsal and peripheral ciliated stripes (Figure 1A), which contain ciliated sensory neurons involved in detecting water flow (rheosensation) We took a candidate-based approach to gain a deeper mechanistic insight into how the sensory stripe cells are specified from a heterogeneous soxB1-2+progenitor pool. POU4 genes are involved in the development of sensory organs detecting mechanical stimulation in divergent organisms (Manley and Ladher, 2008; Zhao et al., 2020). Thus, we investigated the expression and function of S. mediterranea pou4 genes. The planarian genome encodes two POU4 homologs, pou4-1 (also referred to as pou4-like) and pou4-2 (Önal et al., 2012a). In previous work, we found that pou4-1 (pou4-like) is expressed in the planarian CNS (Cowles et al., 2014). In contrast, analysis of pou4-2 using whole-mount in situ hybridization (WISH) showed expression localized in the dorsal head tip and dorsal and peripheral ciliated stripes of intact planarians (Figure 1A-B), a stereotyped pattern common to rheosensory genes (Ross et al., 2018). In addition, pou4-2 expression was detected in cells dispersed throughout the body in a subepidermal punctate pattern and in the cephalic ganglia and ventral nerve cords (Figure 1B). Because POU4 genes have been implicated as terminal selectors in widely divergent organisms (Leyva-Diaz et al., 2020), we examined the expression pattern in regeneration blastemas. During the first 24 hours of regeneration, pou4-2 expression was absent from the blastema. We first detected clear expression on day 3 of regeneration (Figure 1C). The patterning of pou4-2 expression in the blastema at day 3 was less organized and not confined to its normal spatial location, with pou4-2+ cells sparsely scattered throughout the regeneration blastema. During days 4 and 5, pou4-2+ cells began to repopulate the stereotypical stripe expression pattern, and by day 7, proper patterning was restored. In planarians, the regeneration blastema is populated by post-mitotic progenitors (Reddien, 2018). The delayed reestablishment of pou4-2 expression suggests that it functions in the late stages of cell differentiation during regeneration.

soxB1-2 is expressed in and regulates transcription in dorsal and peripheral ciliated stripe neurons as well as in other neural populations, and the epidermis of S. mediterranea (King et al., 2024; Ross et al., 2018). Therefore, we searched the existing scRNA-seq data from the entire body and brain (Fincher et al., 2018) to examine the potential relationship between soxB1-2 and pou4-2. First, we extracted 1427 putative neuronal cells expressing soxB1-2. We resolved 19 distinct soxB1-2+ neuronal clusters (Supplementary File S1 and Figure 2A), of which cluster 8 was marked by pou4-2. The presence of synapsin and synaptogamin (neural markers) in cluster 8 indicated that these cells are neurons (Figure 2B). We combed through the dataset to identify genes that are differentially expressed in the pou4-2-enriched cluster (Figure 2C; Supplementary File S2); highly enriched genes included pkd1L-2 and hmcn-1-L, which are highly enriched in the planarian rheosensory organ (Ross et al., 2018). Because pkd1L-2 and hmcn-1-L expression requires soxB1-2 activity, we hypothesized that soxB1-2 regulates pou4-2 expression. Thus, we treated planarians with soxB1-2 or pou4-2 dsRNA (the RNAi treatment scheme is depicted in Supplemental Figure 1A) and processed them for WISH. We observed a significant reduction in mechanosensory neuron-patterned pou4-2 expression in soxB1-2(RNAi) planarians, whereas pou4-2 expression in the central nervous system remained unaltered (Figure 2D). Conversely, soxB1-2 expression was downregulated in mechanosensory neuron-patterned areas important for rheosensation (Supplementary Figure S2B). However, other areas enriched with soxB1-2 expression, such as the auricles - anteriorly-positioned lateral flaps involved in chemotaxis (Almazan et al., 2021), the pharynx-an organ serving as the entrance and exit to the digestive system (Ishii, 1962), and the epidermis, were unaffected by pou4-2 RNAi. We conclude that soxB1-2 positively regulates pou4-2 expression, specifically in mechanosensory neurons, but not in other cell types.

Smed-pou4-2 is positively regulated by soxB1-2.

(A) UMAP representation of soxB1-2+ neuronal subclusters (denoted by numbers) inferred from a scRNA-seq dataset. (B) pou4-2, synapsin, and synaptogamin are all enriched in the soxB1-2+ cluster 8. (C) Heatmap of genes examined in this study demonstrates their differential expression in pou4-2+ cells. (D) soxB1-2 RNAi caused a reduction in sensory neuron-patterned pou4-2 expression in the dorsal ciliated stripe (dcs) and the peripheral ciliated stripe (pcs) but did not affect ventral nerve cord expression (vnc). Anterior to the top. Scale bars = 200 μm; n ≥ 3 worms tested, with all samples displaying similar expression patterns.

In many organisms, pou4 and the proneural atonal genes are part of the same gene regulatory network (Leyva-Diaz et al., 2020). In Ciona intestinalis, atonal and pou4 are part of a regulatory cascade downstream of Notch signaling that generates sensory neurons (Tang et al., 2013). In mice, atoh1 is required for differentiation of multiple mechanosensory neuron types and stimulates expression of pou4f3 to promote hair cell fate (Yu et al., 2021), and overexpression of pou4f3 together with atoh1 and gfi1 in mouse embryonic stem cells can induce inner ear hair cell differentiation in vitro (Costa et al., 2015). There are three atonal homologs in the planarian genome, but none appear to operate in the same regulatory network as pou4-2. atoh-1 is expressed in a discrete neuronal population in the cephalic ganglia, while atoh8-1 and atoh8-2 are expressed in stem cells in the mesenchyme, and all of them are involved in the regeneration of the nervous system (Cowles et al., 2013). pou4-2 expression was unaffected after RNAi inhibition of all atonal genes in regenerated planarians (Supplementary Figure S2A); conversely, the expression of atonal genes was unaffected in pou4-2 RNAi-treated regenerates (Supplementary Figure S2B). Thus, the functional relationship between pou4 and atonal function observed in other animals does not appear to be conserved in planarians.

Smed-pou4-2 regulates genes involved in sensory neuron terminal fate

Based on known roles of Pou4 genes, we hypothesized that pou4-2 function might be required for cell differentiation and function. To test our hypothesis, we performed RNAi of pou4-2 and pinpointed time points wherein the pou4-2 transcripts were robustly downregulated and subsequently performed whole-animal RNA-seq on day 12 of the RNAi knockdown for pou4-2(RNAi) and control animals (see Methods). Analysis of the resulting data revealed downregulation of putative pou4-2 target genes (Supplementary File S3). Because Pou4 genes are predicted to function as transcriptional activators, we focused further analyses on the downregulated gene set. pou4-2 RNAi RNA-seq uncovered 72 significantly downregulated genes (Figure 3 and Supplementary File S3). GO analysis of the pou4-2-downregulated gene set revealed significant enrichment in ‘Mechanosensation’ (Supplementary File S4), including previously characterized genes we assessed to have roles in planarian mechanosensory modalities like vibration sensation and rheosensation, such as the polycystic kidney disease gene homologs pkd1L-2 and pkd2L-1 genes (Ross et al., 2024; Ross et al., 2018). This result is consistent with prior observations of the pou4-2 RNAi phenotype in scRNA-seq studies and studies on the role of Notch signaling in planarians (Elliott, 2016). The discrete expression and pou4-2 RNA-seq dataset motivated us to characterize the regulatory role of this transcription factor in planarian sensory neuron function and regeneration.

Identification of genes regulated by Smed-pou4-2 using RNA-seq.

Volcano plot of differentially expressed genes after pou4-2 RNAi treatment in intact planarians.

The mechanosensory neurons in the rheosensory organ are distinguished by the expression of multiple sensory neural function genes and consist of at least two distinct populations, marked by the expression of terminal markers polycystic kidney disease 1 like-2 (pkd1L-2) and hemicentin-1-like (hmcn-1-L) (Ross et al., 2018). The evolutionarily conserved PKD1L-2 is a cation channel pore component required for mechanosensation and cilia function (Patel, 2015); pkd1L-2(RNAi) planarians exhibit prominent rheosensory defects (Ross et al., 2024; Ross et al., 2018). On the other hand, hmcn-1-L is an extracellular matrix component involved in anchoring mechanosensory neurons to the epidermis (Vogel and Hedgecock, 2001); no detectable sensory defects are detected in hmcn-1-L(RNAi) planarians (Ross et al., 2018). Consistent with the scRNA-seq data, we found that a subset of pou4-2+cells co-expressed pkd1L-2 or hmcn-1-L, representing two unique populations (Figure 4A). We consistently observed variable expression levels; some cells showed high expression of pou4-2 and low expression of terminal markers (arrows in Figure 4A), while others showed lower expression of pou4-2 but high expression of terminal markers (arrowheads in Figure 4A), and some had high expression of pou4-2 and the terminal markers (white dashed box in Figure 4A). In addition, we observed that certain pou4-2+ cells were negative for pkd1L-2 or hmcn-1-L, suggesting that they may represent undifferentiated progenitors. We reasoned that variable pou4-2 expression might be due to pou4-2 transcripts initially appearing in the progenitor cells to activate transcription of terminal markers and persisting at low levels in terminally differentiated cells.

pou4-2 is required for expression of pkd1L-2 and hmcn-1-L.

(A) Double-fluorescence in situ hybridization to visualize expression of pou4-2 with pkdL1-2 or hmcn-1-L in sensory neurons; 74.8% and 28.4% of pou4-2+ cells were pkd1L-2+ and hmcn-1-L+, respectively. (B) Planarians were fed twice per week for 4 weeks, amputated pre-pharyngeally one day after the last feeding, and allowed to regenerate for 10 days before fixation. Whole-mount in situ hybridization showed pkd1L-2 and hmcn-1-L expression were drastically decreased in pou4-2(RNAi) worms. Blue arrows denote expression in the dorsal and peripheral ciliated stripes (dcs and pcs, respectively). Note that some pkd1L-2 and hmcn-1-L expression was detectable in regenerates (red arrows). Regions denoted by a numbered dashed box indicate that hmcn-1-L expression was unaffected by pou4-2 RNAi, which is shown in zoomed-in insets. Anterior is to the top. Scale bars = 200 μm; n ≥ 3 worms tested, with all samples displaying similar expression patterns.

To test whether this latter population of pou4-2+ cells constitutes sensory neuron progenitors, wild-type planarians were X-ray-treated with ∼100 Gy, a dose reported to deplete early progenitors after 24 hours and late progenitors within 7 days (Eisenhoffer et al., 2008). We performed a WISH analysis every 12 hours post-irradiation and quantified the number of pou4-2+/pkd1L-2-/hmcn-1-L- versus the number of pou4-2+ cells co-labeled with a single fluorophore mix of pkd1L-2 and hmcn-1-L riboprobes (Supplementary Figure 3A). By 5.5 days post-irradiation (dpi), we noticed an obvious decrease in pou4-2+ cells negative for both terminal markers pkd1L-2 and hmcn-1-L compared to the number of cells positive for pou4-2 and either pkd1L-2 or hmcn-1-L (Supplementary Figure 3B-C). WISH analysis was performed at various time points on irradiated animals to compare pou4-2 expression with prog-1 and agat-1 expression, the latter function as early and late progenitors, respectively, in the epidermal differential lineage maturation trajectory (Eisenhoffer et al., 2008). The temporal downregulation of pou4-2+ cells is analogous to that of agat-1+ cells. In contrast, piwi-1+ neoblasts and prog-1 early progenitors were almost entirely depleted by 2-3 dpi (Supplemental Figure 3D), which supports a role for pou4-2+ cells as late progenitors and the source of terminally differentiated mechanosensory neurons. Accordingly, we reasoned that variable pou4-2 expression may be due to the transcripts initially appearing in the progenitor cells to activate transcription of terminal markers and persisting at low levels in terminally differentiated cells.

Next, we asked if pou4-2 function is required to maintain and regenerate pkd1L-2 and hmcn-1-L expression in the dorsal and peripheral ciliated stripes. We conducted WISH on intact and regenerated planarians treated with dsRNA over a time course (Supplemental Figure 1A). Like pou4-2, in control intact or regenerated planarians pkd1L-2 and hmcn-1-L were expressed in the head tip, dorsal ciliated stripe, and dorsal and ventral peripheral stripes (controls in Figure 4B). However, in pou4-2(RNAi) animals, the expression of the marker genes was notably lost, especially in the uninjured worms. Interestingly, in the regenerates, minimal pkd1L-2 and hmcn-1-L expression was observed in the regeneration blastema. We also detected lower levels of expression of hmcn-1-L at scattered locations near peripheral stripes that were unaffected by pou4-2 RNAi (Figure 4B, dashed red boxes). It is possible that a longer pou4-2 RNAi treatment might be necessary to ablate expression in those cells. We conclude that pou4-2 is required for pkd1L-2 and hmcn-1-L expression in the most prominent dorsal and peripheral ciliated stripe cells.

Smed-pou4-2 regulates genes involved in ciliated cell structure organization, cell adhesion, and nervous system development

To further investigate the role of pou4-2 in regulating the differentiation of mechanosensory neurons, we selected eight additional genes from either the RNA-seq pou4-2 RNAi- downregulated genes or the S. mediterranea single-cell RNA-seq database (Figure 2-3; see Supplementary File S2 for the list of genes). Four of the genes, cadherin-23, Smed-Eph1, lipoxygenase homology domain-1, and unconventional myosin VIIA, are homologous to human proteins critical to the normal function of hair cells of the inner ear sensory epithelium, and their mutations lead to sensorineural hearing loss (described below). Using WISH and RNAi analysis, we determined their spatial expression patterns and whether they were downregulated by pou4-2 and soxB1-2 (Figure 5A-B). The expression patterns of calmodulin-2 (calm-2), lipoxygenase homology domain-1 (loxhd-1), and dd_28678 were exclusively confined to the stereotypical mechanosensory neuron pattern in the head tip, body periphery, and dorsal ciliated stripe and completely depleted in pou4-2(RNAi) and soxB1-2(RNAi) planarians (Figure 5A). pou4-2 RNAi-mediated loss of calm2 expression is consistent with observations reported by King et al. (2024). Calmodulins are important for ion channel activity and signal transduction, and human calm2 mutations are associated with delayed neurodevelopment and epilepsy (Crotti et al., 2013). loxhd1 is predicted to encode a highly conserved stereociliary protein involved in hair cell function, and mutated human LOXHD1 causes DFNB77, a form of progressive hearing loss (Grillet et al., 2009). NOP2/Sun RNA methyltransferase family member 7 (nsun-7) was expressed in fewer cells in the mechanosensory neuron pattern overall, but an additional subset of nsun-7+ cells present in the optic cups was unaffected by pou4-2 and soxB1-2 inhibition, in contrast to clearly reduced nsun-7 expression in sensory mechanosensory neuron-rich areas (Figure 5A). In humans, nsun7 activity is required for proper flagella movement and sperm motility, and mutations result in male infertility (Khosronezhad et al., 2015). In control and pou4-2(RNAi) planarians, expression of the monocarboxylate transporter-encoding solute carrier family 16 member 24 (slc16a-24) was detected in the auricles. Auricular slc16a-24 expression was downregulated in soxB1-2(RNAi) but not pou4-2(RNAi) planarians, while expression in the mechanosensory neurons important for rheosensation was downregulated in both (Figure 5A).

Expression analysis of genes co-expressed in Smed-pou4-2+ cells.

(A) WISH images of genes predominantly expressed in the dorsal and peripheral ciliated stripes. WISH post-RNAi revealed reduced expression of mechanosensory neuron-patterned genes (labeled on the left) after soxB1-2 and pou4-2 RNAi (labeled on the top). loxhd-1 was also expressed in a punctate pattern (black arrowheads) that appeared largely unaffected by pou4-2 RNAi. The RNAi treatments did not affect nsun-7 expression in the photoreceptors (blue asterisks). (B) In situ hybridization images from whole-mount and cross-sections of genes expressed in mechanosensory neurons and other cell types. Note reduced expression of mechanosensory neuron-patterned genes after soxB1-2 and pou4-2 RNAi. The red arrowheads highlight the tip cell expression unaffected by soxB1-2 and pou4-2 knockdown in ephA4. The insets show the corresponding cross-section of the worm. Anterior to the left. Blue arrows mark ciliated stripe cell regions. Dashed boxes denote cross-section regions. Abbreviations: cephalic ganglia (cg), dorsal ciliated stripe (dcs), dorsal and ventral peripheral stripes (pcs), epidermis (ep), ventral nerve cords (vnc). Scale bars = 200 µm for intact animals and 100 µm for cross sections; n ≥ 3 worms tested with all samples displaying similar expression patterns.

Other genes we chose to analyze were not exclusively expressed in the ciliated stripes (Figure 5B). Expression of cadherin-23 (cdh23) was highest in the photoreceptors, and low expression was detected in the epidermis. While cdh23 expression was downregulated in the ciliated stripes, cdh23+ cells in the epidermis and photoreceptors were unchanged after pou4-2 and soxB1-2 inhibition. Human CDH23 is expressed in the sensory epithelium of the inner ear, where it is involved in maintaining the stereocilium organization of hair cells required for sound perception and equilibrioception (Kazmierczak et al., 2007), and CDH23 mutations are known to cause hereditary hearing loss (Woo et al., 2014). Smed-Eph1 (Eph1), encoding an ephrin receptor homolog, was also among the differentially expressed genes in soxB1-2+/pou4-2+neurons (Figure 2C). The role of Ephrin signaling in axon guidance is well-established and highly conserved; in mammals, the binding of ligand efnb2 to receptor EphA4 is critical to the differentiation and patterning of hair and support cells on the cochlear sensory epithelium (Defourny et al., 2019) and for targeting and innervating auditory projections to hair cells (Defourny et al., 2013). Consequently, EphA4 mutations and failure to bind ligands are associated with sensorineural hearing loss in humans (Levy et al., 2018). While mechanosensory neuronal patterned expression of Eph1 was downregulated after pou4-2 and soxB1-2 inhibition, low expression in the brain branches of the ventral cephalic ganglia persisted (Figure 4B). Eph1 expression in the anterior-most region of the head tip was downregulated in soxB1-2(RNAi) but was unaffected in pou4-2(RNAi) planarians (Figure 4B). gelsolin-2 (glsn-2) was highly expressed in the sensory neuron pattern and in the epidermis, where it appeared most abundantly expressed near the body periphery and weakly expressed in the medial dorsal surface (Figure 5A). Gelsolins’ roles in nervous system development (Mazur et al., 2016) and as modulators of ciliogenesis and cilium length are evolutionarily conserved (Kim et al., 2010).

Sensory neuronal glsn-2 expression was downregulated in pou4-2(RNAi) and soxB1-2(RNAi) planarians, and epidermal expression was also downregulated in soxB1-2(RNAi) planarians, potentially because of the role of soxB1-2 in ciliated epidermal cell maintenance (Ross et al., 2018).

In the dorsal ciliated stripe, body periphery, and head tip, low expression of unconventional myosin VIIA (myo7a) was detected and depleted in pou4-2(RNAi) and soxB1-2(RNAi) planarians (Figure 5B). Additionally, myo7a was highly expressed in the photoreceptors, and this expression remained in pou4-2(RNAi) and soxB1-2(RNAi) planarians. The low expression of myo7a detected beneath the epidermis was also unaffected by pou4-2 and soxB1-2 inhibition. Human myo7a is important in stereocilium organization, differentiation, and signal transduction of inner ear hair cells (Jaijo et al., 2007). Defective myo7a and cdh23, to a lesser extent, cause Usher Syndrome Type 1B (USH1B), which is characterized by deafness and reduced vestibular function (Roux et al., 2006). neuropeptide precursor-3 (npp-3) was highly expressed in the cephalic ganglia, ventral nerve cords, pharynx, and at lower levels in the parenchyma. A small subset of npp-3+ cells in the dorsal ciliated stripe was depleted in pou4-2(RNAi) and soxB1-2(RNAi) planarians (Figure 5B). Despite the presence of pou4-2+ cells in the ventral nerve cords (Figure 1B), pou4-2 inhibition appeared not to affect npp-3 expression in that region (Figure 5B).

Expression of Smed-pou4-2 is required for mechanosensory neuron regeneration and function

Given that pou4-2 expression is decreased in soxB1-2(RNAi) planarians (Figure 2D) and that pou4-2(RNAi) planarians have decreased expression of genes related to mechanosensory neuron and cilia function (Figure 4-5), we reasoned that subsets of pou4-2+ cells are terminally differentiated ciliated sensory neurons. Moreover, the requirement of soxB1-2 for maintaining ciliated epidermal and sensory neuron populations supports the idea that pou4-2 functions as a terminal selector in planarians. To assess the role of pou4-2 in the dorsal and peripheral ciliated stripes, we first immunostained control and RNAi-treated planarians with anti-Acetylated-Tubulin to mark cilia. Compared to the controls, pou4-2(RNAi) planarians showed decreased cilia labeling along the dorsal ciliated stripe, while ciliated lawns on the dorsal and ventral surfaces remained unchanged (Figure 6A-B). To investigate the role of pou4-2 in mechanosensory function, we used a semi-automated behavioral assay to evaluate the mechanosensory response to vibration stimulation (Ross et al., 2024) (Figure 6C). Knocking down pou4-2 caused a significant attenuation of mechanosensory responses compared to control animals (Figure 6D). Moreover, the vibration stimuli produced significantly reduced responses from the regenerated RNAi animals (Figure 6E). Thus, we conclude pou4-2 is downstream of soxB1-2 and is necessary for maintaining and regenerating ciliated mechanosensory neurons in the rheosensory organ (Figure 6F).

pou4-2 expression is required for mechanosensory neuron regeneration and function.

(A) Acetylated-tubulin staining revealed decreased cilia labeling along the dorsal ciliated stripe after pou4-2 RNAi. Anterior is to the top. Scale bars = 200 μm, n = 4 worms stained for each of the control and experimental groups. (B) Higher magnification of acetylated tubulin staining of control and pou4-2 RNAi animals. Scale bars = 25 μm. (C) Significant reductions in vibration sensation and rheosensation were observed in pou4-2 RNAi worms. These reduced behaviors following stimulation were consistent in intact and regenerates. Data in C are represented as mean ± SD, and n > 25 worms for each experimental group. ****p < 0.0001, Student’s t-test.

Discussion

Smed-pou4-2 plays a key role in the regulation of sensory system differentiation

The interplay between lineage-specifying transcription factors and their respective gene regulatory networks coordinates precise developmental processes and stem cell fate decisions. POU4 transcription factors are conserved terminal selectors of sensory neuron fate, but the role of Pou4 has not been extensively characterized in regeneration. This study’s objective was to elucidate the role of Smed-pou4-2 (pou4-2) in planarian sensory neuron regeneration and to identify gene regulatory network components responsible for maintaining mechanosensory neuron function. pou4-2+ cells include ciliated mechanosensory neurons that allow planarians to detect water currents and vibrations (rheosensation), a function similar to POU4F3 in hair cells of the mammalian inner ear sensory epithelium responsible for auditory perception and equilibrioception. We showed that pou4-2 regulates the expression of genes whose homologs are involved in stereocilium organization, cell adhesion, and nervous system development in other organisms, and their human homologs are implicated in sensorineural hearing loss. These findings shed light on the molecular mechanisms underlying planarian regeneration and provide insight into the conserved role of POU4 transcription factors in sensory neuron development across divergent species.

Analysis of pou4-2 RNAi data revealed differentially expressed genes with known roles in mechanosensory functions, such as loxhd-1, cdh23, and myo7a. Mutations in these genes can cause a loss of mechanosensation/transduction. For example, a recent study has demonstrated that loxhd-1 mutation in the inner hair cell does not affect the structural integrity of the hair cell bundle but rather prevents the activation of MET (mechanoelectrical transducer) channels, thereby contributing to progressive hearing loss in mice (Trouillet et al., 2021). Other interesting candidates included a planarian homolog of Ephrin receptors, Smed-Eph1 (Eph1), which was also expressed in soxB1-2+/pou4-2+ neurons and was downstream of these two transcription factors. Little is known about the role of Ephrin signaling in planarian regeneration. We discovered that Eph1 is required for patterning of mechanosensory neurons in S. mediterranea (McCubbin, 2022; Warner, 2024), which led us to examine in detail how Ephrin signaling genes contribute to neural patterning in planarians (unpublished observations). Thus, as we demonstrated with soxB1-2 (Ross et al., 2024; Ross et al., 2018), planarians are also useful to analyze pou4-2-regulated genes and their roles in cell differentiation or mechanosensation. It will be important to perform a comparative analysis of Pou4-regulated genes gleaned from other animals, like sea anemones or vertebrates like birds and fish, which can regenerate hair cells or the lateral line, respectively (Chen et al., 2019; Tourniere et al., 2020; Zhao et al., 2020).

However, we found that our RNA-seq experimental design was limited in detecting pou4-2- regulated transcripts due to the limitations of systemic RNAi and collecting RNA from how animals. Before leveraging existing transcriptomic data on Pou4 homologs or hair cell or lateral line regeneration (Jiang et al., 2014; Ku et al., 2014; Tourniere et al., 2020) for orthologous comparisons in other species to test to what extent the Pou4 gene regulatory kernel is conserved among these widely divergent animals, we have designed new experiments to enrich for pou4-2-expressing planarian tissues and have performed RNA-seq experiments producing a larger differentially expressed gene set (unpublished). In addition, ATAC-seq experiments could be performed to examine how pou4-2 activity affects chromatin architecture. These future genomic experiments should build upon this work and improve the resolution of the pou4-2 gene regulatory network implicated in sensory neuron regeneration.

pou4-2 function could encompass additional roles other than the ones identified in this study. In addition to the pkd1L-2+ and hmcn-1-L+ populations present in the rheosensory organ, there is a population of pou4-2+ cells in the ventral nerve cords and cephalic ganglia, which are not regulated by SoxB1-2 (Figure 1A). Although the ventral nerve cords are populated by npp-3+ cells, Pou4-2 activity does not regulate npp-3 expression in this region as it does in the rheosensory organ (Figure 5B), and it has not been confirmed whether any cells in the ventral nerve cords are pou4-2+/npp-3+ co-expressing cells. This work did not elucidate the function of this pou4-2+ central nervous system cell population, but it may be possible to mine new scRNA-seq datasets to uncover transcription factors, such as in King et al. (2024), to predict the identities of pou4-2+ cells negative for expression of sensory neuron markers like pkd1L-2 and hmcn-1-L. It was surprising to find that the relationship between pou4-2 and atonal is not conserved in planarians, but the expression patterns of planarian atonal genes indicated that they represent completely different cell populations from pou4-2-regulated mechanosensory neurons. Although many pou4-2-regulated genes are expressed in additional tissues separate from the rheosensory organ, Pou4-2 predominantly appears to regulate gene expression in areas enriched with mechanosensory cells, consistent with observations in pou4-2(RNAi) planarians in other studies (Elliott, 2016; King et al., 2024).

Smed-pou4-2 is required for regeneration of mechanosensory function

RNAi of pou4-2 eliminated the expression of terminally differentiated pkd1L-2 and hmcn-1-L mechanosensory neurons in intact animals, whereas pou4-2(RNAi) regenerates had a low, dispersed expression of these genes at the head tips and peripheral ciliated stripes (Figure 4B). We speculate that following injury, Pou4-2+ post-mitotic progenitors or committed cells can still differentiate into sensory neurons in response to injury and polarity cues. This is consistent with previous observations in animals treated with hydroxyurea (HU) to block stem cell progression through S-phase. In these experiments, Smed-APC-1(RNAi) and Smed-ptc(RNAi) HU-treated animals were still able to regenerate neurons (Evans et al., 2011). It is also consistent with the hypothesis that many planarian stem cells are already specialized (Raz et al., 2021) and may have been unaffected by our RNAi treatment scheme. Nevertheless, both pou4-2(RNAi) intact animals and regenerates exhibited a significant reduction in mechanosensation (i.e., rheosensation and mechanosensation) relative to controls (Figure 6D-E). Moreover, the late re-expression of pou4-2 during regeneration and the faster depletion rate of pou4-2+ cells that were negative for markers of terminal differentiation, pkd1L-2h and hmcn-1-L, compared to loss of pou4-2+ cells positive for the same terminal markers in irradiated planarians (Supplementary Figure S3) points to the pou4-2+ cells as progenitors responsible for generating terminally differentiated pkd1L-2+ and hmcn-1-L+ cells.

pou4-2+ cells in the dorsal head tip and peripheral and dorsal ciliated stripes (the planarian rheosensory organ) are downstream of SoxB1-2 activity, and RNAi and WISH experiments demonstrated that pou4-2 expression is necessary for maintaining the functional properties of mechanosensory neurons (Figure 2D-E). However, the planarian rheosensory organ is composed of both ciliated mechanosensory neuronal and epidermal populations. Since soxB1-2 is a pioneer transcription factor of the ectodermal lineage, RNAi of soxB1-2 resulted in the loss of both ciliated sensory neuronal and epidermal populations in the rheosensory organ of planarian; thus, the dorsal ciliated stripe observed in acetylated tubulin labeling disappeared entirely (Ross et al., 2018). In contrast, in pou4-2(RNAi) animals, only mechanosensory neuronal populations were removed from the dorsal ciliated stripe, but modest labeling of ciliated epidermal populations persisted in the region (Figure 6A-B). Altogether, it can be concluded that pou4-2+ specifically marks subsets of planarian mechanosensory neuron populations and does not have a role in the regulation of the epidermal cells of the rheosensory organ, unlike SoxB1-2, which plays an earlier and broader regulatory role.

Concluding Remarks

Despite molecular evidence indicating planarians possess ciliated mechanoreceptors sharing homology with mechanoreceptor function and development in other organisms, we have yet to resolve the cellular morphologies of the collection of cells comprising the ciliated stripes.

Although it remains uncertain whether ciliated mechanoreceptors are products of convergent evolution or share a common cellular ancestry (Manley and Ladher, 2008), the role of Pou4 appears to represent a critical component of an adaptable gene regulatory network that has been co-opted to manufacture mechanoreceptors in distinct cell types. Secondly, we have yet to fully explore how pou4+ progenitors are born in response to local cues. Studies have implicated Notch signaling as the likely culprit (Elliott, 2016), and recent studies in Schmidtea mediterranea have elegantly demonstrated that Notch signaling plays a role in patterning neurons and glial cells (Scimone et al., 2025). This work demonstrates that pou4-2 plays a key role in ciliated mechanoreceptor maintenance, regeneration, and function in S. mediterranea. Future studies will help to identify additional target pou4-2 genes and further contribute to understanding how Pou4-2 regulates chromatin to facilitate the regeneration of mechanosensing cells in planarians.

eLife Assessment

This is a valuable study that explores the role of the conserved transcription factor POU4-2 in the maintenance, regeneration, and function of planarian mechanosensory neurons. The authors provide solid evidence provided by gene expression and functional studies to demonstrate that POU4-2 is required for the maintenance and regeneration of functional mechanosensory neurons in planarians. Furthermore, the authors identify conserved genes associated with human auditory and rheosensory neurons as potential targets of this transcription factor.

Reviewer #1 (Public review):

Summary:

In this manuscript, the authors explore the role of the conserved transcription factor POU4-2 in planarian maintenance and regeneration of mechanosensory neurons. The authors explore the role of this transcription factor and identify potential targets of this transcription factor. Importantly, many genes discovered in this work are deeply conserved, with roles in mechanosensation and hearing, indicating that planarians may be a useful model with which to study the roles of these key molecules. This work is important within the field of regenerative neurobiology, but also impactful for those studying the evolution of the machinery that is important for human hearing.

Strengths:

The paper is rigorous and thorough, with convincing support for the conclusions of the work.

Weaknesses:

Weaknesses are relatively minor and could be addressed with additional experiments or changes in writing.

Reviewer #2 (Public review):

Summary:

In this manuscript, the authors investigate the role of the transcription factor Smed-pou4-2 in the maintenance, regeneration, and function of mechanosensory neurons in the freshwater planarian Schmidtea mediterranea. First, they characterize the expression of pou4-2 in mechanosensory neurons during both homeostasis and regeneration, and examine how its expression is affected by the knockdown of soxB1, 2, a previously identified transcription factor essential for the maintenance and regeneration of these neurons. Second, the authors assess whether pou4-2 is functionally required for the maintenance and regeneration of mechanosensory neurons.

Strengths:

The study provides some new insights into the regulatory role of pou4-2 in the differentiation, maintenance, and regeneration of ciliated mechanosensory neurons in planarians.

Weaknesses:

The overall scope is relatively limited. The manuscript lacks clear organization, and many of the conclusions would benefit from additional experiments and more rigorous quantification to enhance their strength and impact.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public review):

Summary:

In this manuscript, the authors explore the role of the conserved transcription factor POU4-2 in planarian maintenance and regeneration of mechanosensory neurons. The authors explore the role of this transcription factor and identify potential targets of this transcription factor. Importantly, many genes discovered in this work are deeply conserved, with roles in mechanosensation and hearing, indicating that planarians may be a useful model with which to study the roles of these key molecules. This work is important within the field of regenerative neurobiology, but also impactful for those studying the evolution of the machinery that is important for human hearing.

Strengths:

The paper is rigorous and thorough, with convincing support for the conclusions of the work.

Weaknesses:

Weaknesses are relatively minor and could be addressed with additional experiments or changes in writing.

Reviewer #2 (Public review):

Summary:

In this manuscript, the authors investigate the role of the transcription factor Smed-pou4-2 in the maintenance, regeneration, and function of mechanosensory neurons in the freshwater planarian Schmidtea mediterranea. First, they characterize the expression of pou4-2 in mechanosensory neurons during both homeostasis and regeneration, and examine how its expression is affected by the knockdown of soxB1, 2, a previously identified transcription factor essential for the maintenance and regeneration of these neurons. Second, the authors assess whether pou4-2 is functionally required for the maintenance and regeneration of mechanosensory neurons.

Strengths:

The study provides some new insights into the regulatory role of pou4-2 in the differentiation, maintenance, and regeneration of ciliated mechanosensory neurons in planarians.

Weaknesses:

The overall scope is relatively limited. The manuscript lacks clear organization, and many of the conclusions would benefit from additional experiments and more rigorous quantification to enhance their strength and impact.

Reviewing Editor Comments:

(1) Quantification of pou4-2(+) cells that express (or do not express) hmcn-1-L and/or pkd1L-2(-) is a common suggestion amongst reviewers. It is recognized that Ross et al. (2018) showed that pkd1L-2 and hmcn-1L expression is detected in separate cells by double FISH, and the analysis presented in Supplementary Figure S3 is helpful in showing that some cells expressing pou4-2 (magenta) are not labeled by the combined signal of pkd1L-2 and hmcn-1-L riboprobes (green). However, I am not sure that we can conclude that pkd1L-2 and hmcn-1-L are effectively detected when riboprobes are combined in the analysis. Therefore, quantification of labeled cells as proposed by Reviewers 1 and 2 would help.

Combining riboprobes is a standard approach in the field, and we chose this method as a direct way to determine which cells lack expression of both genes. We agree that providing the raw quantification data would be helpful for readers, and we included this data in Supplementary File S7; the file contains the quantification information for this dFISH experiment represented in Supplementary Figure 3.

(2) It may be helpful to comment on changes (or lack of changes) in atoh gene RNA levels in RNAseq analyses of pou4-2 animals. As mentioned by one of the reviewers, in situs that don't show signal are inconclusive in this regard.

We fully agree with both reviewers. Two of the planarian atonal homologs are difficult to detect and produce background signals, which we attempted and previously reported in Cowles et al. Development (2013). We conceived performing reciprocal RNAi/in situ experiments, born out of curiosity given the reported role of atonal in the pou4 cascade in other organisms. However, these exploratory experiments lacked a strong rationale for inclusion, particularly given that pou4-2 and the atonal homologs do not share expression patterns, co-expression, or differential expression in our RNA-seq dataset. Therefore, we decided to omit the atonal in situs following pou4-2 RNAi. We retained the experiments showing that knockdown of the atonal genes does not show robust effects on the mechanosensory neuron pattern, as expected. We thank the reviewing editor and reviewers for pinpointing the concern. We agree that additional experiments, such as qPCR experiments, would be needed. We reasoned that while these additional experiments could be informative, they are unlikely to alter the key conclusions of this study substantially.

(3) There seem to be typos at bottom of Figure 10 and top of page 11 when referencing to Figure 4B (should be to 5B instead): "While mechanosensory neuronal patterned expression of Eph1 was downregulated after pou4-2 and soxB1-2 inhibition, low expression in the brain branches of the ventral cephalic ganglia persisted (Figure 4B)."

Thank you! We have fixed those.

(4) Typo (page 13; kernel?): "...to test to what extent the Pou4 gene regulatory kernel is conserved among these widely divergent animals."

Regulatory kernels are defined as the minimal sets of interacting genes that drive developmental processes and are the core circuits within a gene regulatory network, but we recognize that this might not be as well known, so we have changed the term to “network” for clarity.

Reviewer #1 (Recommendations for the authors):

(1) The authors indicate that they are interested in finding out whether POU4-2 is important in the creation of mechanosensory neurons in adulthood as well as in embryogenesis (in other words, whether the mechanism is "reused during adult tissue maintenance and regeneration"). The manuscript clearly shows that planarian POU4 -2 is important in adult neurogenesis in planarians, but there is no evidence presented to show that this is a recapitulation of embryogenesis. Is pou4-2 expressed in the planarian embryo? This might be possible to examine by ISH or through the evaluation of sequencing data that already exists in the literature.

We agree that these statements should be precise. We have clarified when we make comparisons to the role of Pou4 in sensory system development in other organisms versus its role in the adult planarian. We examined its expression using the existing database of embryonic gene expression. Thanks for hinting at this idea. We performed BLAST in Planosphere (Davies et al., 2017) to cross-reference our clone matching dd_Smed_v6_30562_0_1, which is identical to SMED30002016. The embryonic gene expression for SMED30002016 indicates this gene is expressed at the expected stages given prior knowledge of the timing of organ development in Schmidtea mediterranea (a positive trend begins at Stage 5, with a marked increase by Stage 6 that remains comparable to the asexual expression levels shown). We thank the reviewer for pointing out this oversight. We have incorporated this result in the paper as a Supplementary Figure and discuss how we can only speculate that it has a similar role as we detect in the adult asexual worms.

(2) Can it be determined whether the punctate pou4-2+ cells outside of the stripes are progenitors or other neural cell types? Are there pou4-2+ neurons that are not mechanosensory cell types? Could there be other roles for POU4-2 in the neurogenesis of other cell types? It might help to show percentages of overlap in Figure 4A and discuss whether the two populations add up to 100% of cells.

These are good questions that arise in part from other statements that need clarification in the text (pointed out by Reviewer 2). We think some of the dorsal pou4-2+ might represent progenitor cells undergoing terminal differentiation (see Supplementary Figure 4). We attempted BrdU pulse chase experiments but were not successful in consistently detecting pou4-2 at sufficient levels with our protocol. In response to this helpful comment, we have included this question as a future direction in the revised Discussion. Finally, we have edited our description of the expression pattern. We already pointed out that there are other cells on the ventral side that are not affected when soxB1-2 is knocked down. We attempted to resolve the potential identity of those cells working with existing scRNA-seq data in collaboration with colleagues, but their low abundance made it difficult to distinguish other populations. While we acknowledge this interesting possibility, we have chosen to focus this report on the role of pou4-2 downstream of soxB1-2, as this represents the most well-supported aspect of the dataset and was positively highlighted by both the reviewer and editor.

(3) The authors discuss many genes from their analysis that play conserved roles in mechanosensation and hearing. Were there any conserved genes that came up in the analysis of pou4-2(RNAi) planarians that have not yet been studied in human hearing and neurodevelopment? I am wondering the extent to which planarians could be used as a discovery system for mechanosensory neuron function and development, and discussion of this point might increase the impact of this paper or provide critical rationale for expanding work on planarian mechanosensation.

Indeed, we agree that planarians could be used to identify conserved genes with roles in mechanosensation and have included this point in the Discussion. In this study, we have focused on demonstrating the conservation of gene regulation. While this study was initially based on a graduate thesis project, we have since generated a more comprehensive dataset from isolated heads, which we are currently analyzing. This has been emphasized in the revised Discussion.

Minor:

(1) For Figure 6E, the authors could consider showing data along a negative axis to indicate a decrease in length in response to vibration and to more clearly show that this decrease doesn't occur as strongly after pou4-2(RNAi).

We displayed this behavior as the percent change, as this is a standard way to represent this data. As the percent change is a positive value, we represent the data as these positive values.

(2) The authors should consider quantifying the decrease of pou4-2 mRNA after atonal(RNAi) conditions, either by RT-qPCR or cell quantification. Visually, the signal in the stripes after atoh8-2(RNAi) seems lower, particularly in the tail. The punctate pattern outside the stripes may also be decreased after atoh8-1(RNAi). But quantification might strengthen the argument.

We agree with the reviewer and acknowledge that we should have been more cautious in interpreting these results. Those two genes are difficult to detect and did not show specific patterns in Cowles et al. (2013). The reviewer is correct that additional experiments are necessary before reaching conclusions, but we do not think as discussed earlier we do not think new experiments would provide insights for the major conclusions. These experiments were exploratory in nature and tangential to our main conclusions, especially in the absence of reciprocal evidence (e.g., shared expression patterns, co-expression, or differential expression in our RNA-seq data. Therefore, we decided to eliminate the atonal in situs following pou4-2 RNAi.

Reviewer #2 (Recommendations for the authors):

A. Expression of pou4-2 in ciliated mechanosensory neurons:

(1) The conclusion that pou4-2 is expressed in ciliated mechanosensory neurons is primarily based on co-expression analysis using a published single-cell dataset. Although the authors later show that a subset of pou4-2 cells also express pkd1L-2 (Figure 4A), a known marker of ciliated mechanosensory neurons, this finding is not properly quantified. I recommend moving Figure 4A to earlier in the manuscript (e.g., to Figure 2) and expanding the analysis to include additional known markers of this cell type. Proper quantification of the extent of co-localization is necessary to support the claim robustly.

As pointed out by the reviewer, there is substantive evidence from our lab and other reports. King et al. also showed pou4-2 and pkd1L-2 ‘regulation’ by their scRNA-seq data, and this function is conserved in the acoel Hofstenia miamia (Hulett et al., PNAS 2024 ). Our analysis shows convincing co-localization by scRNA-seq and expression of soxB1-2 and neural markers in the respective populations. Furthermore, we included colocalization of pou4-2 with mechanosensory genes using fluorescence in situ hybridization (Figure 3B, Supplementary Figure 4, and Supplementary File S7). We are confident the data conclusively show pou4-2 regulates pkd1L-2 expression in a subset of mechanosensory neurons. Given the strength of existing observations and previously published data, we believe that additional staining experiments are not essential to support this conclusion.

(2) There appears to be a conceptual inconsistency in the interpretation of pou4-2 expression dynamics. On one hand, the authors suggest that delayed pou4-2 expression indicates a role in late-stage differentiation (p.6). On the other hand, they propose that pou4-2 may be expressed in undifferentiated progenitors to initiate downstream transcriptional programs (p.8). These interpretations should be reconciled. Additionally, claims regarding pou4-2 expression in progenitor populations should be supported by co-localization with established stem cell or progenitor markers, rather than inferred from signal intensity alone.

This is an excellent point, and we agree with the reviewer that this section requires editing. As described in response to Reviewer 1, we attempted BrdU pulse chase experiments but were not successful in consistently detecting pou4-2 at sufficient levels with our protocol. Furthermore, we could not obtain strong signals in double labeling experiments in pou4-2 in situs combined with piwi-1 or PIWI-1 antibodies. We will include those experiments as a future direction and amend our conclusions accordingly.

(3) The expression pattern shown in Figure 1B raises questions about the precise anatomical localization of pou4-2 cells. It is unclear whether these cells reside in the subepidermal plexus or the deeper submuscular plexus, which represent distinct neuronal layers (Ross et al., 2017). The observed signals near the ventral nerve cords could suggest submuscular localization. To clarify this, higher-resolution imaging and co-staining with region-specific neural markers are recommended.

In Ross et al. (2018), we showed that the pkd1L-2+ cells are located submuscularly. The pkd1L-2 cells express pou4-2, thus the pou4-2+ cells are located in the same location. Based on co-expression data and co-expression with PKD genes, we are confident it is submuscular.

B. The functional requirements of pou4-2 in the maintenance of mechanosensory neurons:

(1) To evaluate the functional role of pou4-2 in maintaining mechanosensory neurons, the authors performed whole-animal RNA-seq on pou4-2(RNAi) and control animals, identifying a significant downregulation of genes associated with mechanosensory neuron expression. However, the presentation of these findings is fragmented across Figures 3, 4, and 5. I recommend consolidating the RNA-seq results (Figure 3) and the subsequent validation of downregulated genes (Figures 4 and 5) into a single, cohesive figure. This would improve the logical flow and clarity of the manuscript.

As suggested by the reviewer, we have combined Figures 3 and 4 (new Figure 3), which we believe improves the flow. We decided to keep Figure 5 (new Figure 4) as a standalone because it focuses on the characterization of new genes revealed by RNAseq and scRNA-seq data mining that were not previously reported in Ross et al. 2018 and

1.

(2) In pou4-2(RNAi) animals, pkd1L-2 expression appears to be entirely lost, while hmcn-1-L shows faint expression in scattered peripheral regions. The authors suggest that an extended RNAi treatment might be necessary to fully eliminate hmcn-1-L expression. However, an alternative explanation is that pou4-2 is not essential for maintaining all hmcn-1-L cells, particularly if pou4-2 expression does not fully overlap with that of hmcn-1-L. This possibility should be acknowledged and discussed.

We agree and have acknowledged this point in the revised text.

(3) On page 9, the section title claims that "Smed-pou4-2 regulates genes involved in ciliated cell structure organization, cell adhesion, and nervous system development." While some differentially expressed genes are indeed annotated with these functions based on homology, the manuscript does not provide experimental evidence supporting their roles in these biological processes in planarians. The title should be revised to avoid overstatement, and the limitations of extrapolating a function solely from gene annotation should be acknowledged.

Excellent point. We have edited the text to indicate that the genes were annotated or implicated.

(4) The cilia staining presented in Figure 6B to support the claim that pou4-2 is required for ciliated cell structure organization is unconvincing. Improved imaging and more targeted analysis (e.g., co-labeling with mechanosensory markers) are needed to support this conclusion.

We have addressed this concern by adjusting the language to be more precise and indicate that the stereotypical banded pattern is disrupted with decreased cilia labeling along the dorsal ciliated stripe. Indeed, our conclusion overstated the observations made with the staining and imaging resolution. Thank you.

C. The functional requirements of pou4-2 in the regeneration of mechanosensory neurons:

To evaluate the role of pou4-2 in the regeneration of mechanosensory neurons, the authors performed amputations on pou4-2(RNAi) and control(RNAi) animals and assessed the expression of mechanosensory markers (pkd1L-2, hmcn-1-L) alongside a functional assay. However, the results shown in Figure 4B indicate the presence of numerous pkd1L-2 and hmcn-1-L cells in the blastema of pou4-2(RNAi) animals. This observation raises the possibility that pou4-2 may not be essential for the regeneration of these mechanosensory neurons. The authors should address this alternative interpretation.

Our interpretation is that there were very few cells expressing the markers compared to controls. The pattern was predominantly lost, which is consistent with other experiments shown in the paper. However, we have added the additional caveat suggested by the reviewer.

Minor points:

(1) On p.8, the authors wrote "every 12 hours post-irradiation". However, this is not consistent with the figure, which only shows 0, 3, 4, 4.5, 5, and 5.5 dpi.

We corrected this. Thank you for catching the mistake!

(2) On p.12, the authors wrote "Analysis of pou4-2 RNAi data revealed differentially expressed genes with known roles in mechanosensory functions, such as loxhd-1, cdh23, and myo7a. Mutations in these genes can cause a loss of mechanosensation/transduction". This is misleading because, to my knowledge, the role of these genes in planarians is unknown. If the authors meant other model systems, they should clearly state this in the text and include proper references.

The reviewer is correct that we are referencing findings from other organisms. We have clarified this point in the revised text. The appropriate references were included and cited in the first version.

(3) On p.7, the authors wrote, "conversely, the expression of atonal genes was unaffected in pou4-2 RNAi-treated regenerates (Supplementary Figure S2B)". However, it is unclear whether the Atoh8-1 and Atoh8-2 signals are real, as the quality of the in situ results is too low to distinguish between real signals and background noise/non-specific staining.

This valid concern was addressed in our response to Reviewer 1. We have adjusted the figure and the text accordingly.

(4) On p.6 the authors wrote "pinpointed time points wherein the pou4-2 transcripts were robustly downregulated". However, the current version of the manuscript does not provide data explaining why Pou4-2 transcripts are robustly downregulated on day 12.

Yes, we determined the appropriate time points using qPCR for all sample extractions. As an example, see the figure for qPCR validation at day 12 showing that pou4-2 and pkd1L2 are down.

In this graph, samples labeled “G” represent four biological controls of gfp(RNAi) control animals, and samples labeled “P” represent four biological controls of pou4-2(RNAi)animals at day 12 in the RNAi protocol.

(5) On p.13, the authors wrote "collecting RNA from how animals." Is this a typo?

Thanks for catching the typo. It should read “whole” animals. We have corrected this.

(6) On p.14, the authors wrote "but the expression patterns of planarian atonal genes indicated that they represent completely different cell populations from pou4-2-regulated mechanosensory neurons". However, this is unclear from the images, as the in situ staining of Atoh8-1 and Atoh82 are potentially failed stainings.

We agree. We have edited accordingly.

Version 2

Abstract

POU4 homologs are involved in the development of sensory cell types across diverse species, including cnidarians, ascidians, and mammals. Whether these developmental regulators are redeployed during adult tissue maintenance and regeneration is an open question in regenerative biology. Here, we investigate the role of the Schmidtea mediterranea BRN3/POU4 homolog, Smed-pou4-2 (pou4-2), in the regeneration of mechanosensory neurons. We find that pou4-2 is regulated by the SoxB1 homolog, soxB1-2, and is expressed in a distinct population of ciliated sensory cells that detect water flow. Transcriptomic analysis of pou4-2-deficient planarians reveals enrichment for conserved genes associated with human auditory and vestibular function, suggesting that planarian rheosensory neurons share molecular features with mammalian inner ear hair cells. Expression of these conserved genes is significantly reduced following RNAi-mediated knockdown of pou4-2. To determine whether these transcriptional changes have functional consequences for mechanosensory neuron identity or behavior, we assessed the impact of pou4-2 knockdown on sensory function. pou4-2 RNAi results in impaired mechanosensation in both uninjured and regenerating planarians. Together with the loss of terminal differentiation markers in mechanosensory neurons, these findings identify Smed-pou4-2 as a key regulator of mechanosensory neuron identity in planarians and support the idea that conserved sensory specification programs are redeployed during adult tissue regeneration.

Introduction

Most animals, including mammals, have a limited capacity for neuronal regeneration. In contrast, organisms like fish and salamanders can effectively regenerate neurons, and some invertebrates are capable of dramatic whole-body regeneration. The freshwater planarian Schmidtea mediterranea is among a handful of research organisms capable of restoring virtually any lost or damaged tissue and can regenerate entire animals from small body fragments (Goldstein and Srivastava, 2022; Ivankovic et al., 2019). S. mediterranea possesses a population of adult pluripotent stem cells called neoblasts, which proliferate and differentiate to replace all missing tissues (Baguñà, 2012; Newmark and Sánchez Alvarado, 2002; Reddien, 2018). This stem cell population is postulated to include a heterogeneous pluripotent pool poised to acquire lineage-specific cell fates as needed (Raz et al., 2021). One of the extraordinary properties of planarians is the capacity for constant neuronal turnover and regeneration of neuronal cell types, many of which are conserved with vertebrates (Brown and Pearson, 2017; Ross et al., 2017). Despite recent advances, much remains unknown about the molecular basis of neurogenesis and the signals that regulate neuronal turnover (Lee, 2023).

Previous studies found that soxB1-2, a mammalian Sox1/2/3 homolog, regulates the regeneration of ectodermal cell type subsets in planarians, including many uncharacterized sensory neurons (Ross et al., 2018). One prominent population of soxB1-2-regulated sensory cells, organized in a striking dorsal stripe pattern, functions in mechanosensation and is marked by polycystic kidney disease-like homologs (Ross et al., 2024; Ross et al., 2018) (Fig. 1A). In this study, we focused on identifying mechanisms downstream of soxB1-2 that contribute to the specification of mechanosensory cells in the dorsal ciliated stripe.

Smed-pou4-2 is expressed in the ciliated stripes.

(A) Schematic of dorsal and peripheral sensory cell patterns implicated in mechanosensation. (B) Whole-mount in situ hybridization (WISH) of pou4-2 reveals stereotyped mechanosensory neuron expression in the dorsal head tip, body periphery, dorsal ciliated stripe (dcs), and ventral nerve cords (vnc). The dashed line indicates the cross-section plane shown below. Scale bar = 200 μm. (C) Regeneration time-course showing reappearance of pou4-2 expression in the blastema beginning at day 3 post-amputation (asterisk), with re-establishment of dorsal stripe expression by day 7. Blue arrows mark the reappearing dorsal ciliated stripe pattern. Anterior is up. Scale bars = 200 µm; n ≥ 3.

POU transcription factor family genes play key roles in the development and function of many neuronal subtypes. To date, dozens of POU genes have been identified in vertebrates and invertebrates, and their roles in the differentiation and survival of diverse neuronal subtypes have been characterized (Leyva-Diaz et al., 2020). In many species, Brn3/POU4 transcription factors play important roles in specifying and maintaining the identities of various cell populations in the developing peripheral sensory nervous system. A notable example is Nematostella vectensis NvPOU4, which is required to maintain and differentiate cnidocytes, a population of mechanosensing cells exclusive to the phylum Cnidaria (Tourniere et al., 2020). The homologous role of NvPOU4 in cnidarians suggests that the functional role of pou4 is ancient and conserved across distantly related phyla. Additionally, pou4 is part of a proneural regulatory cascade that produces epidermal sensory neurons in Ciona intestinalis; induction of ectopic pou4 expression in the developing epidermis of Ciona larvae converts epidermal cells to sensory neurons, resulting in a striking hyper-ciliated phenotype (Chen et al., 2011).

In mice, Pou4f3 is expressed in the inner ear sensory epithelia during embryonic development and is required for the survival of vestibular hair cells of the auditory system (Erkman et al., 1996; Xiang et al., 1997) - its targeted deletion results in impaired hearing and balance. Hair cells of the inner ear are crosslinked by stereocilia on their apical ends that function as mechanosensors, converting vibration-induced mechanical force into signals carried by auditory nerve fibers to the central nervous system (Goutman et al., 2015). Although a small number of hair cells differentiate in Pou4f3-/- mice, their failure to form stereociliary bundles leads to apoptosis (Xiang et al., 1998). Thus, Pou4 has conserved roles in the differentiation, maintenance, and survival of ciliated mechanosensory neurons. Unlike birds and fish, mammals lack the ability to regenerate hair cells after they are lost, resulting in permanent deafness (Edge and Chen, 2008). Recent studies show that POU4 can be used as a reprogramming co-factor to restore hair cells in mammals (Chen et al., 2021; Iyer et al., 2022). However, whether developmental regulators like POU4 play similar roles in adult tissue maintenance and regeneration remains to be fully resolved (Seifert et al., 2023).

In S. mediterranea, a search for candidate planarian OCT4 homologs, a gatekeeper of pluripotency also known as POU5F1 in humans (Zeineddine et al., 2014), revealed six genes containing a POU-specific domain and a POU-homeodomain, and the two genes most similar to hPOU4F3 were named Smed-pou4-1 and Smed-pou4-2 (Önal et al., 2012b) (referred to as pou4-1 and pou4-2 hereon). pou4-1 (also referred to as pou4-like and pou4-like-1) was identified as downstream of COE (Cowles et al., 2014), a transcription factor required for neurogenesis widely conserved across metazoans (Demilly et al., 2011), and is responsible for maintaining proper neuronal architecture in the cephalic ganglia as well as photoreceptor pigmentation (Cowles et al., 2014). More recently, our lab and others observed robust sensory defects in pou4-2(RNAi) planarians (Elliott, 2016; McCubbin, 2022; Wang, 2019). Under normal conditions, S. mediterranea worms display a stereotyped behavior by shortening their bodies in response to vibrations and water currents across their dorsal side (rheosensation). We found that pou4-2(RNAi) planarians failed to react to this sensory input, suggesting a critical role for pou4-2 in mechanosensory neuron function (Elliott, 2016; Wang, 2019); however, its role in regenerative neurogenesis is not well understood. Here, we examined the function of pou4-2 in mechanosensory neuron regeneration.

In this study, we mapped the expression of pou4-2 and assessed its function through RNAi and RNA-seq. Loss of pou4-2 expression coincides with loss of mechanosensation, which is not restored in regenerated pou4-2(RNAi) planarians. Analysis of the pou4-2+ cell gene expression profile uncovered that many genes regulated by Pou4-2 activity are necessary for proper mechanosensory neuron function and are homologs of human genes involved in hair cell function and auditory perception. In many organisms, the proneural atonal genes function in the same gene regulatory network as pou4 (Leyva-Diaz et al., 2020). However, this relationship does not appear to be conserved in planarians. Our findings suggest that pou4-2 participates in a regulatory cascade involved in the specification of distinct sensory neuron populations. This study demonstrates that pou4-2 plays a key regulatory role in the differentiation, maintenance, and regeneration of ciliated mechanosensory neurons in planarians.

Results

Smed-pou4-2 is expressed in planarian mechanosensory neurons

Our previous work demonstrated a key role for soxB1-2 in the differentiation and function of sensory neuron subclasses in the planarian Schmidtea mediterranea (Ross et al., 2018). A subset of soxB1-2-regulated genes are abundantly expressed in a discrete pattern called the dorsal and peripheral ciliated stripes (Figure 1A), which contain ciliated sensory neurons involved in detecting water flow (rheosensation). We took a candidate-based approach to gain mechanistic insight into how the sensory stripe cells are specified from a heterogeneous soxB1-2+ progenitor pool. POU4 genes are involved in the development of sensory organs detecting mechanical stimulation in divergent organisms (Manley and Ladher, 2008; Zhao et al., 2020).

Thus, we investigated the expression and function of S. mediterranea pou4 genes. The planarian genome encodes two POU4 homologs, pou4-1 (also referred to as pou4-like) and pou4-2 (Önal et al., 2012a). In previous work, we found that pou4-1 (pou4-like) is expressed in the planarian CNS (Cowles et al., 2014). In contrast, analysis of pou4-2 using whole-mount in situ hybridization (WISH) showed expression localized in the dorsal head tip and dorsal and peripheral ciliated stripes of intact planarians (Figure 1A-B), a stereotyped pattern common to rheosensory genes (Ross et al., 2018). In addition, pou4-2 expression was detected in cells dispersed throughout the body in a subepidermal punctate pattern and in the cephalic ganglia and ventral nerve cords (Figure 1B). Because POU4 genes have been implicated as terminal selectors in diverse organisms (Leyva-Diaz et al., 2020), we examined the expression pattern of pou4-2 in regeneration blastemas to determine whether its activation coincides with late differentiation stages. During the first 24 hours of regeneration, pou4-2 expression was absent from the blastema. We first detected clear expression on day 3 of regeneration (Figure 1C). The patterning of pou4-2 expression in the blastema at day 3 was less organized and not confined to its normal spatial location, with pou4-2+ cells sparsely scattered throughout the regeneration blastema. During days 4 and 5, pou4-2+ cells began to repopulate the stereotypical stripe expression pattern, and by day 7, proper patterning was restored. In planarians, the regeneration blastema is populated by post-mitotic progenitors (Reddien, 2018). The delayed re-establishment of pou4-2 expression suggests that it is required in the later stages of cell differentiation in regeneration. Interestingly, embryonic expression of pou4-2 is predominantly detected during Stages 5 to 6, which coincide with organogenesis and nervous system development in S. mediterranea (Supplementary Figure S1). While the functional role of pou4-2 in embryogenesis remains unknown, these data suggest that its temporal dynamics are consistent with a role in late-stage differentiation, as observed in adult regeneration (Davies et al., 2017).

soxB1-2 is expressed in and regulates transcription in dorsal and peripheral ciliated stripe neurons as well as in other neural populations, and the epidermis of S. mediterranea (King et al., 2024; Ross et al., 2018). Therefore, we searched the existing scRNA-seq data from the entire body and brain (Fincher et al., 2018) to examine the potential relationship between soxB1-2 and pou4-2. First, we extracted 1427 putative neuronal cells expressing soxB1-2. We resolved 19 distinct soxB1-2+ neuronal clusters (Supplementary File S1 and Figure 2A), of which cluster 8 was marked by pou4-2. The presence of synapsin and synaptogamin (neural markers) in cluster 8 indicated that these cells are neurons (Figure 2B). We combed through the dataset to identify genes that are differentially expressed in the pou4-2+ cluster (Figure 2C; Supplementary File S2); highly enriched genes included pkd1L-2 and hmcn-1-L, which are highly enriched in the planarian rheosensory organ (Ross et al., 2018). Because pkd1L-2 and hmcn-1-L expression requires soxB1-2 activity, we hypothesized that soxB1-2 regulates pou4-2 expression. Thus, we treated planarians with soxB1-2 or pou4-2 dsRNA (the RNAi treatment scheme is depicted in Supplemental Figure 2A) and processed them for WISH. We observed a significant reduction in mechanosensory neuron-patterned pou4-2 expression in soxB1-2(RNAi) planarians, whereas pou4-2 expression in the central nervous system remained unaltered (Figure 2D). Conversely, soxB1-2 expression was downregulated in mechanosensory neuron-patterned areas important for rheosensation (Supplementary Figure S2B). However, other areas enriched with soxB1-2 expression, such as the auricles - anteriorly-positioned lateral flaps involved in chemotaxis (Almazan et al., 2021), the pharynx - an organ serving as the entrance and exit to the digestive system (Ishii, 1962), and the epidermis, were unaffected by pou4-2 RNAi. Together, these results support a model in which soxB1-2 positively regulates pou4-2 expression specifically in mechanosensory neurons, without affecting pou4-2 expression in other neural or epidermal populations.

Smed-pou4-2 is positively regulated by soxB1-2.

(A) UMAP of soxB1-2⁺ neuronal subclusters from scRNA-seq data. (B) pou4-2, synapsin, and synaptogamin are enriched in cluster 8. (C) Heatmap of genes examined in this study demonstrates their differential expression in the pou4-2+ cell cluster. (D) soxB1-2 RNAi reduces pou4-2 expression in the mechanosensory dorsal and peripheral ciliated stripes (dcs and pcs), but not in the ventral nerve cords (vnc) or cephalic ganglia (cg). Scale bars = 200 μm; n ≥ 3 worms tested, with all samples displaying similar expression patterns.

In many organisms, pou4 and the proneural atonal genes are part of the same gene regulatory network (Leyva-Diaz et al., 2020). In Ciona intestinalis, atonal and pou4 are part of a regulatory cascade downstream of Notch signaling that generates sensory neurons (Tang et al., 2013). In mice, atoh1 is required for differentiation of multiple mechanosensory neuron types and stimulates expression of pou4f3 to promote hair cell fate (Yu et al., 2021), and overexpression of pou4f3 together with atoh1 and gfi1 in mouse embryonic stem cells can induce inner ear hair cell differentiation in vitro (Costa et al., 2015). There are three atonal homologs in the planarian genome, but none appear to operate in the same regulatory network as pou4-2. atoh-1 is expressed in a discrete neuronal population in the cephalic ganglia, while atoh8-1 and atoh8-2 are expressed in stem cells in the mesenchyme, and all of them are involved in the regeneration of the nervous system (Cowles et al., 2013). pou4-2 expression was unaffected after RNAi inhibition of all atonal genes in regenerated planarians (Supplementary Figure S3). Thus, the functional relationship between pou4 and atonal observed in other animals does not appear to be conserved in planarians, based on current expression data and RNAi analyses.

Smed-pou4-2 regulates genes involved in sensory neuron terminal fate

Based on known roles of Pou4 genes, we hypothesized that pou4-2 is required for sensory neuron differentiation and function. To test our hypothesis, we performed RNAi of pou4-2 and pinpointed time points wherein the pou4-2 transcripts were robustly downregulated (not shown) and subsequently performed whole-animal RNA-seq on day 12 of the RNAi knockdown for pou4-2(RNAi) and control animals (see Methods). Analysis of the resulting data revealed downregulation of putative pou4-2 target genes (Supplementary File S3). Because Pou4 genes are predicted to function as transcriptional activators, we focused further analyses on the downregulated gene set. pou4-2 RNAi RNA-seq uncovered 72 significantly downregulated genes (FC ≥ 1.4, adjusted p-value < 0.1,) (Figure 3A and Supplementary File S3). GO analysis of the pou4-2-downregulated gene set revealed significant enrichment in ‘Mechanosensation’ (Supplementary File S4), including previously characterized genes we assessed to have roles in planarian mechanosensory modalities like vibration sensation and rheosensation, such as the polycystic kidney disease gene homologs pkd1L-2 and pkd2L-1 genes (Ross et al., 2024; Ross et al., 2018). This result is consistent with prior observations of the pou4-2 RNAi phenotype in scRNA-seq studies and studies on the role of Notch signaling in planarians (Elliott, 2016). The discrete expression and pou4-2 RNA-seq dataset motivated us to characterize the regulatory role of this transcription factor in planarian sensory neuron function and regeneration.

Identification of genes regulated by Smed-pou4-2 using RNA-seq.

(A) Volcano plot of genes differentially expressed in pou4-2(RNAi) animals compared to controls (FC ≥ 1.4, p-adj ≤ 0.1). A subset of genes examined in this study is highlighted on the plot and demonstrates significant downregulation. (B) Co-localization analysis by double-fluorescence in situ hybridization reveals that 74.8% of pou4-2⁺ cells co-express pkd1L-2, and 28.4% co-express hmcn-1-L. White boxed cells shown in insets show high pou4-2 and terminal marker expression and are displayed at higher magnification. White arrowheads point to examples where terminal marker gene expression is much brighter than pou4-2 expression. White arrows mark pou4-2+cells with high expression of pou4-2 and low expression of the terminal marker genes. Scale bar = 100 μm. (C) WISH of pkd1L-2 and hmcn-1-L in control and pou4-2(RNAi) 10-day regenerates. Terminal marker expression is strongly reduced in RNAi animals. Numbered red boxes demonstrate a population of scattered hmcn-1-L+ cells that persist following pou4-2 RNAi and are shown in corresponding zoomed-in insets. Blue arrows denote expression in the dorsal and peripheral ciliated stripes (dcs and pcs, respectively). Note that some pkd1L-2 and hmcn-1-L expression was detectable in regenerates (magenta arrows). Anterior is to the top. Scale bars = 200 μm; n ≥ 3 worms tested, with all samples displaying similar expression patterns.

The mechanosensory neurons in the rheosensory organ are distinguished by the expression of multiple sensory neural function genes and consist of at least two distinct populations, marked by the expression of terminal markers polycystic kidney disease 1 like-2 (pkd1L-2) and hemicentin-1-like (hmcn-1-L) (Ross et al., 2018). The evolutionarily conserved PKD1L-2 is a cation channel pore component required for mechanosensation and cilia function (Patel, 2015); pkd1L-2(RNAi) planarians exhibit prominent rheosensory defects (Ross et al., 2024; Ross et al., 2018). On the other hand, hmcn-1-L is an extracellular matrix component involved in anchoring mechanosensory neurons to the epidermis (Vogel and Hedgecock, 2001); no detectable sensory defects are detected in hmcn-1-L(RNAi) planarians (Ross et al., 2018). Consistent with the scRNA-seq data, we found that a subset of pou4-2⁺ cells co-expressed pkd1L-2 (74.8%) or hmcn-1-L (28.4%), representing two distinct mechanosensory neuron subtypes (Figure 3B, Supplementary File S7). We consistently observed variable expression levels; some cells showed high expression of pou4-2 and low expression of terminal markers (arrows in Figure 3B), while others showed lower expression of pou4-2 but high expression of terminal markers (arrowheads in Figure 3B), and some had high expression of pou4-2 and the terminal markers (white dashed box in Figure 3B). We also observed pou4-2⁺ cells lacking detectable expression of either pkd1L-2 or hmcn-1-L, which may represent late-stage progenitors that have not yet initiated terminal marker expression. Thus, the variable pou4-2 and terminal marker expression could be due to pou4-2 transcripts initially appearing at high levels in differentiating progenitor cells (initially terminal marker negative) to activate transcription of terminal markers and then persisting at lower levels in terminally differentiated cells.

Next, we asked if the function of pou4-2 is required to maintain and regenerate pkd1L-2 and hmcn-1-L expression in the dorsal and peripheral ciliated stripes. We conducted WISH on intact and regenerated planarians treated with dsRNA over a time course (Supplemental Figure 1A). Like pou4-2, in control intact or regenerated planarians, pkd1L-2 and hmcn-1-L were expressed in the head tip, dorsal ciliated stripe, and dorsal and ventral peripheral stripes (controls in Figure 3C). In pou4-2(RNAi) animals, expression of both marker genes was strongly reduced, particularly in intact animals. In regenerating animals, expression was detected at lower levels and in scattered patterns. Interestingly, in the regenerates, minimal pkd1L-2 and hmcn-1-L expression was observed in the regeneration blastema. We also detected lower levels of expression of hmcn-1-L at scattered locations near peripheral stripes that were unaffected by pou4-2 RNAi (Figure 3C, dashed red boxes). These results indicate that pou4-2 is necessary for maintaining pkd1L-2 and hmcn-1-L expression in the most prominent dorsal and peripheral ciliated stripe cell populations. The persistence of hmcn-1-L expression in some peripheral cells may reflect partial knockdown efficiency or that pou4-2 function is not required in all hmcn-1-L⁺ cells.

To test whether the population of pou4-2+/terminal marker- cells in dorsal ciliated stripe constitutes sensory neuron progenitors that have yet to express terminal markers in support of our hypothesis above, wild-type planarians were X-ray-treated with ∼100 Gy, a dose reported to deplete early progenitors after 24 hours and late progenitors within 7 days (Eisenhoffer et al., 2008). We performed WISH analysis at 0, 3, 4, 4.5, 5, and 5.5 days post-irradiation (dpi). We quantified pou4-2⁺ cells lacking expression of pkd1L-2 and hmcn-1-L, compared to pou4-2⁺ cells co-labeled using a single fluorophore mix of pkd1L-2 and hmcn-1-L riboprobes (Supplementary Figure 4A). By 5.5 days post-irradiation (dpi), we noticed an obvious decrease in pou4-2+/terminal marker- compared to the number of pou4-2+/terminal marker+ cells (Supplementary Figure 4B-C). The temporal decline in pou4-2⁺ cells closely resembled that of agat-1⁺ late progenitors, suggesting a similar position within the differentiation trajectory. In contrast, piwi-1⁺ neoblasts and prog-1⁺ early progenitors were almost entirely depleted by 2–3 days dpi, supporting the idea that pou4-2 marks a later-stage progenitor population (Supplementary Figure 4D). These data support a model in which pou4-2 expression is established in late-stage progenitor cells prior to the expression of terminal sensory neuron genes and remains expressed at lower levels to maintain expression of sensory function genes in terminally differentiated mechanosensory neurons.

Smed-pou4-2 regulates genes implicated in ciliated cell structure organization, cell adhesion, and nervous system development

To further investigate the role of pou4-2 in regulating the differentiation of mechanosensory neurons, we selected eight additional genes from either the RNA-seq pou4-2 RNAi- downregulated genes or the S. mediterranea single-cell RNA-seq database (Figure 2-3; see Supplementary File S2 for the list of genes). Four of the selected genes, cadherin-23 (cdh-23), Ephrin Receptor 1 (EphR1), lipoxygenase homology domain-1 (loxhd-1), and unconventional myosin VIIA (myo7a), encode proteins homologous to those required for the proper function of inner ear hair cells in humans. Mutations in the human homologs of these genes are associated with sensorineural hearing loss (described below). We used WISH and RNAi analysis to determine their spatial expression patterns and assess whether expression was downregulated by pou4-2 or soxB1-2 knockdown (Figure 4A-B). The expression patterns of calmodulin-2 (calm-2), loxhd-1, and dd_28678 were exclusively confined to the stereotypical mechanosensory neuron pattern in the head tip, body periphery, and dorsal ciliated stripe and completely depleted in pou4-2(RNAi) and soxB1-2(RNAi) planarians (Figure 4A). pou4-2 RNAi-mediated loss of calm2 expression is consistent with observations reported by King et al. (2024).

Expression analysis of genes co-expressed in Smed-pou4-2+ cells.

(A) WISH images of genes predominantly expressed in mechanosensory neuron regions, dorsal and peripheral ciliated stripes (dcs, pcs). WISH post-RNAi revealed reduced expression of mechanosensory neuron-patterned genes (labeled on the left) after soxB1-2 and pou4-2 RNAi (labeled on the top). loxhd-1 was also expressed in a punctate pattern (black arrowheads) that appeared largely unaffected by pou4-2 RNAi. The RNAi treatments did not affect nsun-7 expression in the photoreceptors (blue asterisks). (B) In situ hybridization images from whole-mount and cross-sections of genes expressed in mechanosensory neurons and other cell types. Note reduced expression of genes in the stereotypical mechanosensory neuron regions after soxB1-2 and pou4-2 RNAi. The red arrowheads highlight the head tip expression unaffected by soxB1-2 and pou4-2 knockdown in EphR1-labeled cells. The insets show the corresponding cross-section of the worm. Anterior to the left. Blue arrows mark ciliated stripe cell regions. Dashed boxes denote cross-section regions. Abbreviations: cephalic ganglia (cg), dorsal ciliated stripe (dcs), dorsal and ventral peripheral stripes (pcs), epidermis (ep), ventral nerve cords (vnc). Scale bars = 200 µm for intact animals and 100 µm for cross sections; n ≥ 3 worms tested with all samples displaying similar expression patterns.

Calmodulins are important for ion channel activity and signal transduction, and human calm2 mutations are associated with delayed neurodevelopment and epilepsy (Crotti et al., 2013). loxhd1 is predicted to encode a highly conserved stereociliary protein involved in hair cell function, and mutated human LOXHD1 causes DFNB77, a form of progressive hearing loss (Grillet et al., 2009). NOP2/Sun RNA methyltransferase family member 7 (nsun-7) was expressed in fewer cells in the mechanosensory neuron pattern overall, but an additional subset of nsun-7+ cells present in the optic cups was unaffected by pou4-2 and soxB1-2 inhibition, in contrast to clearly reduced nsun-7 expression in sensory mechanosensory neuron-rich areas (Figure 4A). In humans, nsun7 activity is required for proper flagella movement and sperm motility, and mutations result in male infertility (Khosronezhad et al., 2015). In control and pou4-2(RNAi) planarians, expression of the monocarboxylate transporter-encoding solute carrier family 16 member 24 (slc16a-24) was detected in the auricles. Auricular slc16a-24 expression was downregulated in soxB1-2(RNAi) but not pou4-2(RNAi) planarians, while expression in the mechanosensory neurons important for rheosensation was downregulated in both (Figure 4A).

Other genes we chose to analyze were not exclusively expressed in the ciliated stripes (Figure 4B). Expression of cadherin-23 (cdh23) was highest in the photoreceptors, and low expression was detected in the epidermis. While cdh23 expression was downregulated in the ciliated stripes, cdh23+ cells in the epidermis and photoreceptors were unchanged after pou4-2 and soxB1-2 inhibition. Human CDH23 is expressed in the sensory epithelium of the inner ear, where it is involved in maintaining the stereocilium organization of hair cells required for sound perception and equilibrioception (Kazmierczak et al., 2007), and CDH23 mutations are known to cause hereditary hearing loss (Woo et al., 2014). Smed-EphR1 (EphR1), encoding an ephrin receptor homolog, was also among the differentially expressed genes in soxB1-2+/pou4-2+ neurons (Figure 2C). The role of Ephrin signaling in axon guidance is well-established and highly conserved; in mammals, the binding of ligand efnb2 to receptor EphA4 is critical to the differentiation and patterning of hair and support cells on the cochlear sensory epithelium (Defourny et al., 2019) and for targeting and innervating auditory projections to hair cells (Defourny et al., 2013). In humans, mutations in EphA4 and disruption of ephrin ligand binding are associated with sensorineural hearing loss (Levy et al., 2018). While mechanosensory neuronal patterned expression of EphR1 was downregulated after pou4-2 and soxB1-2 inhibition, low expression in the brain branches of the ventral cephalic ganglia persisted (Figure 4B). EphR1 expression in the anterior-most region of the head tip was downregulated in soxB1-2(RNAi) but was unaffected in pou4-2(RNAi) planarians (Figure 4B). gelsolin-2 (glsn-2) was highly expressed in the sensory neuron pattern and in the epidermis, where it appeared most abundantly expressed near the body periphery and weakly expressed in the medial dorsal surface (Figure 5A). Gelsolins’ roles in nervous system development (Mazur et al., 2016) and as modulators of ciliogenesis and cilium length are evolutionarily conserved (Kim et al., 2010).

pou4-2 expression is required for mechanosensory neuron regeneration and function.

(A) Acetylated-tubulin staining shows decreased cilia signal in the dorsal ciliated stripe of pou4-2(RNAi) animals. Scale bars = 200 μm, n = 4 worms stained for each of the control and experimental groups. (B) Higher magnification confirms stripe reduction; epidermal and ventral cilia are unaffected. Scale bars = 25 μm. (C) Vibration response assay demonstrating body contractions in wild-type animals following tapping stimulus. (D, E) Quantification of vibration response assay data shows significantly reduced contraction responses in both intact (D) and regenerate (E) pou4-2(RNAi) animals. Data in D and E are represented as mean ± SD; n ≥ 25 worms for each experimental group. ****p < 0.0001, Student’s t-test. (F) Model: pou4-2 acts downstream of soxB1-2 in regulating mechanosensory neuron differentiation.

Mechanosensory glsn-2 expression was downregulated in both pou4-2(RNAi) and soxB1-2(RNAi) planarians. Epidermal expression of glsn-2 was also reduced in soxB1-2(RNAi) animals, consistent with its broader role in ciliated epidermal cell maintenance (Ross et al., 2018).

In the dorsal ciliated stripe, body periphery, and head tip, low expression of unconventional myosin VIIA (myo7a) was detected and depleted in pou4-2(RNAi) and soxB1-2(RNAi) planarians (Figure 4B). Additionally, myo7a was highly expressed in the photoreceptors, and this expression remained in pou4-2(RNAi) and soxB1-2(RNAi) planarians. The low expression of myo7a detected beneath the epidermis was also unaffected by pou4-2 and soxB1-2 inhibition. Human myo7a is important in stereocilium organization, differentiation, and signal transduction of inner ear hair cells (Jaijo et al., 2007). Defective myo7a and cdh23, to a lesser extent, cause Usher Syndrome Type 1B (USH1B), which is characterized by deafness and reduced vestibular function (Roux et al., 2006). neuropeptide precursor-3 (npp-3) was highly expressed in the cephalic ganglia, ventral nerve cords, pharynx, and at lower levels in the parenchyma. A small subset of npp-3+ cells in the dorsal ciliated stripe was depleted in pou4-2(RNAi) and soxB1-2(RNAi) planarians (Figure 4B). Despite the presence of pou4-2+ cells in the ventral nerve cords (Figure 1B), pou4-2 inhibition appeared not to affect npp-3 expression in that region (Figure 5B).

Expression of Smed-pou4-2 is required for mechanosensory neuron regeneration and function

Given that pou4-2 expression is reduced in soxB1-2(RNAi) planarians (Figure 2D) and that pou4-2(RNAi) animals show downregulation of mechanosensory and cilia-related genes (Figures 3-4), we reasoned that a subset of pou4-2⁺ cells represents terminally differentiated ciliated sensory neurons. Moreover, the requirement of soxB1-2 for maintaining ciliated epidermal and sensory neuron populations supports the idea that pou4-2 functions as a terminal selector in planarians. To assess the role of pou4-2 in the dorsal and peripheral ciliated stripes, we first immunostained control and RNAi-treated planarians with anti-Acetylated-Tubulin to mark cilia. Compared to the controls, pou4-2(RNAi) planarians showed a disruption of the stereotypical banded pattern with decreased cilia labeling along the dorsal ciliated stripe, while ciliated lawns on the dorsal and ventral surfaces remained unchanged (Figure 5A-B). To investigate the role of pou4-2 in mechanosensory function, we used a semi-automated behavioral assay to evaluate the mechanosensory response to vibration stimulation (Ross et al., 2024) (Figure 5C). Knockdown of pou4-2 led to a significant reduction in vibration-induced contraction behavior in both intact and regenerating animals (Figure 5D–E). Thus, we conclude pou4-2 is downstream of soxB1-2 and is necessary for maintaining and regenerating ciliated mechanosensory neurons in the rheosensory organ (Figure 5F).

Discussion

Smed-pou4-2 plays a key role in the regulation of sensory system differentiation

The interplay between lineage-specifying transcription factors and their respective gene regulatory networks coordinates precise developmental processes and stem cell fate decisions. POU4 transcription factors are conserved terminal selectors of sensory neuron fate, but the role of Pou4 has not been extensively characterized in regeneration. This study’s objective was to elucidate the role of Smed-pou4-2 (pou4-2) in planarian sensory neuron regeneration and to identify gene regulatory network components responsible for maintaining mechanosensory neuron function. pou4-2+ cells include ciliated mechanosensory neurons that allow planarians to detect water currents and vibrations (rheosensation), a function similar to POU4F3 in hair cells of the mammalian inner ear sensory epithelium responsible for auditory perception and equilibrioception. We showed that pou4-2 regulates the expression of genes homologous to those involved in stereocilium organization, cell adhesion, and nervous system development in other organisms. Several of these human homologs are implicated in sensorineural hearing loss. These findings shed light on the molecular mechanisms underlying planarian regeneration and provide insight into the conserved role of POU4 transcription factors in sensory neuron development across divergent species.

Differential expression analysis following pou4-2 knockdown revealed several genes homologous to those with known roles in mechanosensation in other organisms, including loxhd-1, cdh23, and myo7a. In vertebrate systems, mutations in these genes disrupt mechanosensory function and contribute to hearing loss. For example, a recent study has demonstrated that loxhd-1 mutation in the inner hair cell does not affect the structural integrity of the hair cell bundle but rather prevents the activation of MET (mechanoelectrical transducer) channels, thereby contributing to progressive hearing loss in mice (Trouillet et al., 2021). Other interesting candidates included a planarian homolog of Ephrin receptors, EphR1, which was also expressed in soxB1-2+/pou4-2+ neurons and was downstream of these two transcription factors. This observation highlights the potential for planarians to serve as a discovery platform for conserved regulators of sensory neuron patterning. We discovered that EphR1 is required for patterning of mechanosensory neurons in S. mediterranea (McCubbin, 2022; Warner, 2024), which led us to examine in detail how Ephrin signaling genes contribute to neural patterning in planarians (unpublished observations). Thus, as we demonstrated with soxB1-2 (Ross et al., 2024; Ross et al., 2018), planarians are also useful to analyze pou4-2-regulated genes and their roles in cell differentiation or mechanosensation. It will be important to perform a comparative analysis of Pou4-regulated genes gleaned from other animals, like sea anemones or vertebrates like birds and fish, which can regenerate hair cells or the lateral line, respectively (Chen et al., 2019; Tourniere et al., 2020; Zhao et al., 2020). However, we found that our RNA-seq experimental design had limitations in detecting pou4-2-regulated transcripts, likely due to the use of systemic RNAi and bulk RNA extraction from whole animals. Before leveraging existing transcriptomic data on Pou4 homologs or hair cell or lateral line regeneration (Jiang et al., 2014; Ku et al., 2014; Tourniere et al., 2020) for orthologous comparisons in other species to test to what extent the Pou4 gene regulatory network is conserved among these widely divergent animals, we have designed new experiments to enrich for pou4-2-expressing planarian tissues and have performed RNA-seq experiments producing a larger differentially expressed gene set (unpublished). In addition, ATAC-seq experiments could be performed to examine how pou4-2 activity affects chromatin architecture. These future genomic experiments should build upon this work and improve the resolution of the pou4-2 gene regulatory network implicated in sensory neuron regeneration.

pou4-2 function could encompass additional roles other than the ones identified in this study. In addition to the pkd1L-2+ and hmcn-1-L+ populations present in the rheosensory organ, there is a population of pou4-2+ cells in the ventral nerve cords and cephalic ganglia, which are not regulated by SoxB1-2 (Figure 1A). Although the ventral nerve cords are populated by npp-3+ cells, Pou4-2 activity does not regulate npp-3 expression in this region as it does in the rheosensory organ (Figure 4B), and it has not been confirmed whether any cells in the ventral nerve cords are pou4-2+/npp-3+ co-expressing cells. While this study did not define the function of this pou4-2⁺ central nervous system population, it may be possible to mine new scRNA-seq datasets to uncover transcription factors, such as in King et al. (2024), to predict the identities of pou4-2+ cells negative for expression of sensory neuron markers like pkd1L-2 and hmcn-1-L.

Our results indicate that the regulatory relationship between pou4 and atonal observed in other species does not appear to be conserved in planarians. The expression patterns of planarian atonal genes indicated that they represent completely different cell populations from pou4-2-regulated mechanosensory neurons. Although several pou4-2-regulated genes are also expressed outside the rheosensory organ, pou4-2 appears to primarily regulate gene expression in mechanosensory neuron-enriched regions, consistent with prior observations in pou4-2(RNAi) planarians (Elliott, 2016; King et al., 2024).

Smed-pou4-2 is required for the regeneration of mechanosensory function

RNAi of pou4-2 substantially reduced the expression of the terminal mechanosensory markers pkd1L-2 and hmcn-1-L in intact animals, whereas pou4-2(RNAi) regenerate animals had a markedly reduced, dispersed expression of these genes at the head tips and peripheral ciliated stripes (Figure 3B). These findings raise the possibility that following injury, pou4-2⁺ progenitors or committed precursors may still undergo limited differentiation into sensory neurons in response to injury and polarity cues. This is consistent with previous observations in animals treated with hydroxyurea (HU) to block stem cell progression through S-phase. In these experiments, Smed-APC-1(RNAi) and Smed-ptc(RNAi) HU-treated animals were still able to regenerate neurons (Evans et al., 2011). It is also consistent with the hypothesis that many planarian stem cells are already specialized (Raz et al., 2021) and may have been unaffected by our RNAi treatment scheme. Nevertheless, both pou4-2(RNAi) intact animals and regenerates exhibited a significant reduction in mechanosensory responses compared to controls (Figure 5D-E). Moreover, the late re-expression of pou4-2 during regeneration, combined with the more rapid depletion of pou4-2⁺ cells lacking pkd1L-2 and hmcn-1-L expression compared to pou4-2⁺ cells co-expressing these terminal markers in irradiated animals, supports the idea that pou4-2⁺ progenitors give rise to terminally differentiated mechanosensory neurons (Supplementary Figure S4).

pou4-2+ cells in the dorsal head tip and peripheral and dorsal ciliated stripes (the planarian rheosensory organ) are downstream of SoxB1-2 activity, and RNAi and WISH experiments demonstrated that pou4-2 expression is necessary for maintaining the functional properties of mechanosensory neurons (Figure 2D-E). However, the planarian rheosensory organ is composed of both ciliated mechanosensory neuronal and epidermal populations. Since soxB1-2 is a pioneer transcription factor of the ectodermal lineage, RNAi of soxB1-2 resulted in the loss of both ciliated sensory neuronal and epidermal populations in the rheosensory organ of planarian; thus, the dorsal ciliated stripe observed in acetylated tubulin labeling disappeared entirely (Ross et al., 2018). In contrast, in pou4-2(RNAi) animals, labeling of ciliated mechanosensory neurons was markedly reduced, but epidermal ciliated cells within the stripe appeared largely unaffected (Figure 5A–B). These findings support the conclusion that pou4-2⁺ cells represent a subset of mechanosensory neurons within the rheosensory organ and that pou4-2 is not required for the maintenance of ciliated epidermal cells, unlike soxB1-2, which has a broader role in ectodermal lineage regulation.

Concluding Remarks

Despite molecular evidence indicating planarians possess ciliated mechanoreceptors sharing homology with mechanoreceptor function and development in other organisms, we have yet to fully resolve the cellular morphologies of the collection of cells comprising the ciliated stripes. Although it remains uncertain whether ciliated mechanoreceptors are products of convergent evolution or share a common cellular ancestry (Manley and Ladher, 2008), the role of Pou4 appears to represent a critical component of an adaptable gene regulatory network that has been co-opted to manufacture mechanoreceptors in distinct cell types. In addition, the mechanisms that specify pou4⁺ progenitors in response to local cues remain unclear. Studies have implicated Notch signaling as the likely culprit (Elliott, 2016), and recent studies in Schmidtea mediterranea have elegantly demonstrated that Notch signaling plays a role in patterning neurons and glial cells (Scimone et al., 2025). Together, our findings demonstrate that pou4-2 is critical for the maintenance, regeneration, and function of ciliated mechanosensory neurons in S. mediterranea. Future studies aimed at identifying additional pou4-2 target genes and defining how pou4-2 influences chromatin accessibility will provide deeper insight into the transcriptional regulation of mechanosensory regeneration in planarians.

Experimental methods

Planarian culture

Asexual clonal line CIW4 of S. mediterranea was maintained in 1x Montjuïc salts (1.6 mM NaCl, 1.0 mM CaCl2, 1.0 mM MgSO4, 0.1 mM MgCl2, 0.1 mM KCl, and 1.2 mM NaHCO3) in the dark at 20°C and fed weekly with pureed calf liver (Merryman et al., 2018). Planarians 3-5 mm in length were starved for one week before experimentation unless specified otherwise.

Gene identification and cloning

Sequences were obtained from an EST library (Zayas et al., 2005), cloned using gene-specific primers, or synthesized as eBlocks (IDT) and inserted into pPR-T4P (Liu et al., 2013) or pJC53.2 (Collins et al., 2010) vectors through ligation-independent cloning. Primer sequences, eBlock sequences, and EST clone accession numbers are listed in Supplementary File S5.

In situ hybridization

Riboprobes were synthesized using an in vitro transcription reaction from DNA templates with digoxigenin or fluorescein-labeled NTPs, and whole-mount in situ hybridizations were performed as previously described (King and Newmark, 2013) in an InsituPro automated liquid handling robot (CEM Corporation, Matthews, NC). Briefly, samples were incubated with anti-Digoxigenin-AP (1:2000, Roche) for chromogenic detection, and the signals were subsequently developed with NBT/BCIP in AP buffer. For double fluorescent in situ hybridizations (dFISH), samples were incubated for 16 hours at 4°C with anti-DIG-AP and anti-FITC-POD (1:250, Roche). Peroxidase-conjugates were detected with tyramide signal amplification (TSA) as outlined previously in (Brown and Pearson, 2015), and Fast Blue development was utilized for AP-driven reaction detection (Lauter et al., 2011).

scRNA sequencing data analysis

To infer the gene expression profiles of pou4-2+ cells, we analyzed publicly available scRNA-seq data from S. mediterranea [GSE111764] (Fincher et al., 2018). From the whole-body data (50562 cells) and the brain data (7766 cells), we extracted putative neuronal cells, defined as those expressing at least one of the following transcripts: synapsin (dd_Smed_v4_3135_0_1), synaptotagmin (dd_Smed_v4_4222_0_1, dd_Smed_v4_6730_0_1, dd_Smed_v4_6920_0_1), and synaptosome associated protein 25 (dd_Smed_v4_13079_0_1, dd_Smed_v4_13255_0_1, or dd_Smed_v4_3977_0_1). This yielded 20,557 cells of which 1,427 expressed soxB1-2 (dd_Smed_v4_8104_0_1). Expression data were scaled and transformed using standard functions of the R library Seurat (Satija et al., 2015). We performed dimensionality reduction using UMAP and clustering using the Leiden algorithm (Traag et al., 2019). Since the UMAP projection suggested communities of disparate size, we used the Leiden algorithm in two steps – once with a smaller resolution to capture the large-scale structure (10 clusters) and then selectively with a higher resolution to resolve smaller communities, for a total of 19 distinct clusters. We characterized each cluster by finding differentially expressed genes (markers) using the bootstrap method of Pollard and van der Laan (Pollard and Laan, 2005) to calculate the z-score for every gene between cells in each cluster relative to cells outside the cluster (Supplementary File S1). The z-scores are then assessed for significance using the empirical Bayes method of Efron (Efron, 2008). The result is a posterior error probability lfdr assigned to each gene.

RNA interference (RNAi)

Bacterially expressed dsRNA was prepared by cloning gene-specific fragments into pPR-T4P or pJC.53.2 vectors, then transforming them into HT115 E. coli. Briefly, cultures grown overnight in LB with appropriate antibiotics were diluted in 40 mL 2xYT and incubated at 37°C and shaking at 225 rpm. Once cultures reached 0.6-0.8 OD600, 1 mM IPTG was added to induce dsRNA synthesis, and incubation continued for 2 hours. Cultures were then pelleted at 3,000 x g for 10 minutes at 4°C, resuspended in 8 mL LB, and aliquoted into eight microcentrifuge tubes.

Resuspended cultures were pelleted at 11,000 x g for 5 minutes at 4°C, aspirated, and stored at -80°C. For RNAi feeding, bacterial pellets were mixed with liver puree as previously described (Gurley et al., 2008). Animals were fed 8 times over 4 weeks, with gfp dsRNA used as a negative control. For stainings, intact animals were fixed 10 days after the 8th feed, and regenerated animals were pre-pharyngeally amputated one day after the 8th feed and fixed after 10 days of regeneration. RNAi feedings for the behavioral assays were performed using in vitro transcribed dsRNA mixed with pureed liver and agarose, as described in (Ross et al., 2024). For intact animals, assays were performed three days after the 8th feed. For regenerates, animals were amputated 24 hours after the 8th feed and tested for behavioral defects 14-17 days of regeneration. The total numbers of animals for RNAi experiments are summarized in Supplementary File S6.

RNA sequencing

Three biological replicates were obtained, each consisting of four worms of approximately four mm length at the start of the experiment, which were starved for one week prior to the start of RNAi feeding. Worms were fed bacterially expressed dsRNA three times on days 0, 3, and 7, and RNA was extracted and purified on day 12 (Allen et al., 2021). 500 ng of total RNA was used for the RNA-seq library preparation and sequencing at MedGenome, Inc. (Foster City, CA). The Poly-A-containing mRNA molecules were purified using poly-T oligo attached magnetic beads, and then the mRNA was converted to cDNA using Illumina TruSeq stranded mRNA kit (20020595) according to the manufacturer’s protocol. Libraries were sequenced for 100 cycles to a depth of 30 million paired reads using Illumina NovaSeq 6000 (Illumina, San Diego, CA). The following quality control steps were performed on the fastq files: Base quality score distribution, Sequence quality score distribution, Average base content per read, GC distribution in the reads, distribution of over-represented sequences, and adapter trimming.

Based on the quality report of fastq files, sequences were trimmed wherever necessary to retain only high-quality sequences for further analysis. In addition, the low-quality sequence reads are excluded from the analysis. Data quality check was performed using FastQC (v0.11.8). The adapter trimming was performed using the fastq-mcf program (v1.05) and cutadapt (v2.5) (Martin, 2011). Transcriptome alignment was performed using RSEM (version RSEM v1.3.1) (Li and Dewey, 2011) against the dd_Smed_v6 transcriptome (Rozanski et al., 2019) to build Bowtie transcriptome indexes using rsem-prepare-reference, then used rsem-calculate-expression for aligning and expression calculation. Differential expression analysis was performed using DESeq2 (R Bioconductor package) (Anders and Huber, 2010) with default parameters, and then differentially reduced genes were defined as those having a fold-change of <1.4 and p-adjusted value < 0.1 (see Supplementary File S3). The RNA sequencing data have been deposited in NCBI under BioProject accession PRJNA1258257. Gene Ontology (GO) annotation and over-presentation analysis, the unique set of pou4-2(RNAi) downregulated transcripts was compared to the human proteome (BLASTX against the Swiss-Prot Homo sapiens proteome, cutoff e-value < 1e-3). Human UniProt IDs were used for enrichment analysis using Fisher’s Exact tests with FDR multiple test correction (FDR < 0.05) in http://geneontology.org/. GO results are reported in Supplementary File S4.

Immunohistochemistry

Animals were sacrificed in ice-cold 2% HCl for 30 seconds, followed by incubation in Carnoy’s fixative (6 parts ethanol: 3 parts CHCl3: 1 part glacial acetic acid) for 2 hours at 4°C (Forsthoefel et al., 2018), followed by a dehydration step with 100% methanol for 1 hour at 4°C. The animals were bleached overnight under a lamp with 6% H2O2 in methanol and then rehydrated in 75%, 50%, and 25% methanol-PBSTx, followed by two 5-minute PBSTx washes. PBSTb (1% BSA in PBSTx) was used for blocking at room temperature for 2 hours. Primary antibody labeling was carried out with mouse anti-Acetylated Tubulin (Sigma-Aldrich, St. Louis, MO) diluted in PBSTb (1:1000) overnight at 4°C. Six 1-hour PBSTx washes followed by 1 hour of PBSTb blocking were performed before anti-mouse-HRP (1:1000, Cell Signaling) incubation overnight at 4°C. After six 1-hour PBSTx washes, acetylated tubulin was detected through TSA development as described in (Brown and Pearson, 2015) with the following exceptions: no 4-IPBA or dextran sulfate was added to the TSA reaction buffer, Cy3-tyramide was diluted 1:250, and development took place for a total of 20 minutes.

Mechanosensation (vibration) assay

Analysis of the planarian’s ability to detect a vibration stimulus was conducted essentially as described in (Ross et al., 2024). Briefly, groups of five control gfp(RNAi) or pou4-2(RNAi) planarians were added to a 100 × 15 mm petri dish containing 40 ml of 1x Montjuïc salts that was placed inside a dish lid that was mounted to a cold LED lighted board using clear silicone paste and observed until gliding normally. Then, an Arduino-controlled arm delivered five taps at a rate of one tap every 75 msec to the side of the dish. Experimental runs were recorded on a Basler Ace 2 Pro ac1440-220uc camera connected to a PC running Basler’s pylon Viewer 64-bit version 6.3.0 software at a frame rate of 10 frames/sec and a frame size of 1440 × 1080 pixels. The video frames were analyzed in Fiji (ImageJ2 version 2.9.0) (Schindelin et al., 2012), using the line tool to measure the longest pre-stimulus gliding length and the length of the worms following the stimulus. The percent change in length was calculated as [(LengthPrestimulus – LengthPoststimulus)/LengthPrestimulus] × 100. Statistical analysis and graph generation were performed in GraphPad Prism 9 (GraphPad Software, Boston, MA). One-way ANOVA analyses were performed and corrected using Dunnett’s correction. All means were compared to the control group, and statistical significance was accepted at values of p < 0.01.

X-ray irradiation

Starved animals (3–4 mm) were irradiated with 100 Gy of X-rays (130 kV, 5 mA, 8.4 Gy/min) for approximately 12 minutes using a CellRad irradiator (Precision X-Ray, Madison, CT).

Imaging

Brightfield images were acquired with a Leica DFC450 camera and M205 stereomicroscope. Animals processed for fluorescent in situ hybridization and immunohistochemistry were mounted in Vectashield diluted 1:1 in 80% glycerol. Fluorescent images were acquired using a Zeiss AxioZoom equipped with an Apotome using Zen Pro version. High-magnification acetylated tubulin images were acquired using a Zeiss Axio Observer Inverted Microscope equipped with an AV4 Mod Apotome using AxioVision v4.6 (Carl Zeiss Microscopy, LLC, White Plains, NY).

Cell counting and quantification

For co-labeling experiments between pou4-2-regulated genes with pkd1L-2 or hmcn-1-L, maximum intensity projections of stacked fluorescent images were acquired with a depth of 12 µm. For pou4-2 co-labeling with pkd1L-2 and hmcn-1-L in irradiated planarians, maximum intensity projections of stacked fluorescent images were acquired with a depth of 24 µm. The regions quantified spanned the width of the sensory neuron expression pattern in the head tip and 500 µm anterior to the head tip along the length of the dorsal ciliated stripe. Cells were manually counted using Zen Lite v3.3. Cell quantification was represented as the number of positive cells per mm2, and graphs were made using GraphPad Prism (GraphPad Software, Boston, MA). Three to six biological replicates per group were used to quantify co-labeling. Cell quantification details and the total number of cells counted for Figure 3B and Supplementary Figure S4A-B are summarized in Supplementary Files S7 and S8, respectively.